# Expressions and names {#sec-expressions-names}

```{r}
#| include: false
source(here::here("_common.R"))
```

Open R. Type `1 + 1` and press Enter.

```{r}
1 + 1
```

You just wrote a complete R program. It has no variables, no functions, no setup. It is an expression\index{expression}, and R evaluated it and printed the result. Every piece of R code you will ever write works this way: you give R an expression, R gives you back its value.

What makes this work? Three things: R's interactive loop, expressions themselves, and a mechanism for giving names to values so you can use them later.


## The REPL {#sec-repl}

When you open R, you see a prompt: `>`. R is waiting. When you type something and press Enter, four things happen in sequence:

1. R **reads** what you typed.
2. R **evaluates**\index{evaluation} it (figures out the value).
3. R **prints** the result.
4. R **loops** back to step 1 and waits again.

This is called a REPL\index{REPL} (Read-Eval-Print Loop). There is no "compile" step, no "run" button; you type an expression, you see its value immediately.

::: {.callout-note collapse="true"}
## Interpreted or compiled?
In compiled languages such as C, Java, or Go, you write the whole program first and a compiler translates it into machine code before anything runs. R, like Python and JavaScript, is *interpreted*: it reads and evaluates one expression at a time. A compiler can optimize the whole program in advance, so compiled code usually runs faster, but an interpreter lets you see results as you go, which is why interpreted languages dominate data analysis.

The line is blurry in practice. R itself is written in C, and many R functions are thin wrappers around compiled C or Fortran code. When you type `sum(x)`, the arithmetic happens in compiled machine code underneath. You get the interactive interface; the heavy computation still runs at full speed.
:::

Try a few:

```{r}
42
```

```{r}
"hello"
```

```{r}
TRUE
```

```{r}
sqrt(9)
```

Each of these is an expression. `42` is an expression whose value is 42; `"hello"` evaluates to the text string "hello"; `sqrt(9)` calls the function `sqrt` with the argument `9` and returns 3. The `[1]` you see before each result is R telling you "this is the first element of the output." You can ignore it for now; it becomes useful when results have many elements (@sec-vectors).

R evaluates expressions and prints their values, the way Mr. Fold folded `3, 5, 2, 8` down to `18` in @sec-church. This is Church's model of computation, made interactive. But evaluating expressions and forgetting them immediately is like a calculator with no memory. What if you want to keep a result?


### Exercises {.unnumbered}

1. Type `2 + 3 * 4` into R. What do you get? Does R follow the usual order of operations (multiplication before addition)?
2. What happens when you type `"hello" + 1`? Read the error message. What is R telling you?
3. Try `pi`. This is a name that R already knows. What value does it have?


## Names and binding {#sec-names-binding}

You want to compute something, save the result, and use it later. In R, you do this by giving a value a name:

```{r}
x <- 3
```

The arrow `<-` is R's assignment operator\index{assignment}. It binds\index{binding} the name `x` to the value `3`. After this line runs, typing `x` gives you back 3:

```{r}
x
```

You can then use `x` in other expressions:

```{r}
x + 10
```

```{r}
x * x
```

It is tempting to think of `x` as a box that contains 3, and if you have used C or Python, that is exactly the model you carry around: the variable is a container, and assignment puts a value into it. In R, `x <- 3` does not put 3 into a box; it attaches the label `x` to a value that already exists. The value 3 is out there. `x` is just a name for it.

Why does that distinction matter?

```{r}
y <- x
```

Now both `x` and `y` refer to 3. If you later write `x <- 7`, that changes what `x` points to, but `y` still points to 3:

```{r}
x <- 7
x
y
```

`x <- 7` did not change the 3. R made a new value, 7, and moved the label `x` onto it; the 3 is still there, because `y` still points at it. Values in R never change once they exist\index{referential transparency}. When you "change" something, R creates a new value and moves the name, and a value that no name points to any more is reclaimed by R's garbage collector\index{garbage collection}, the mechanism inherited from Lisp (@sec-lisp).

This is why two names can never surprise each other in R. If names were containers, `y <- x` might make two containers share the same contents, and a change through one would show up in the other. With labels on values that never change, there is nothing to share and nothing to corrupt.

::: {.callout-note collapse="true"}
## What this rules out
In C, you learn to watch for *aliasing*: two pointers to the same block of memory, one of them writing while the other reads stale data. In Python, `a = b` for a list means both names see the same mutable object, and modifying through one surprises anyone holding the other. Entire chapters of software engineering textbooks are devoted to defensive copying, synchronization, and ownership protocols, machinery whose sole purpose is to keep shared mutable state from corrupting a program. R removes the condition that makes the checks necessary: values do not change, so that category of bugs\index{side effect} is absent by construction. @sec-stack-and-heap covers how R manages this at the implementation level.
:::

::: callout-tip
## Opinion
Use `<-` for assignment, not `=`. Both work, but `<-` is the convention in R, and for good reason: on the Bell Labs terminals where S was designed, there was a single key that typed `<-` (@sec-s-language). The convention stuck, and almost all R code uses it. Reserve `=` for naming arguments inside function calls, like `log(8, base = 2)`.
:::


### Exercises {.unnumbered}

1. Predict what this code prints, then run it to check:

    ```r
    a <- 10
    b <- a
    a <- 20
    b
    ```

2. What happens if you type a name that doesn't exist, like `zzz`? Read the error message.
3. Names in R are case-sensitive. If you run `X <- 5`, can you get the value back by typing `x`?


## Expressions all the way down {#sec-expressions-all-the-way}

In most programming languages, there are two kinds of code: *expressions* (which produce a value) and *statements* (which do something but don't produce a value). In Python, `1 + 1` is an expression, but `if x > 0: print("yes")` is a statement. You cannot write `result = if x > 0: "yes" else: "no"` because `if` doesn't produce a value.

R has no statements. Everything is an expression, and every expression returns a value.

`if/else` returns a value:

```{r}
x <- 3
result <- if (x > 0) "positive" else "non-positive"
result
```

You can store the result of `if/else` in a name because `if/else` *is* an expression that evaluates to a value, just like `1 + 1`.

A block of code in curly braces `{ }` returns the value of its last expression:

```{r}
result <- {
  a <- 2
  b <- 3
  a + b
}
result
```

The block evaluated three expressions, and the value of the whole block is the value of the last one: `a + b`, which is 5.

Even assignment returns a value (though R hides it). You can reveal it with parentheses:

```{r}
(z <- 42)
```

Without the parentheses, R assigns 42 to `z` and prints nothing. With them, R prints the value that the assignment expression returned.

The rule comes from Church's lambda calculus (@sec-church), which has only expressions. Because every piece of R code produces a value, every piece can be nested inside any other. What happens when you apply that rule to the thing that does the most work in R: functions?


### Exercises {.unnumbered}

1. What does this return? Predict first, then run it: `if (FALSE) 1 else 2`
2. What value does this block produce? `{ 10; 20; 30 }`
3. Can you store the result of `for (i in 1:3) i` in a name? Try it. What happens? (This is one of the very few places where R behaves more like a statement language.)


## Calling functions {#sec-calling-functions}

You have already called a function: `sqrt(9)`. But what is actually happening when R sees those parentheses?

`sqrt` is a name bound to a function, the same way `x` was a name bound to 3. The parentheses `()` tell R to *call* the function, passing `9` as an argument. R evaluates the call and returns the result:

```{r}
sqrt(9)
```

What happens if you type `sqrt` without parentheses?

```{r}
sqrt
```

R prints the function itself. `sqrt` is a value, just like `3` or `"hello"`, and the parentheses are what make R run it. Keep the difference between the function and the call in mind; @sec-functions-are-values is built on it.

Functions can take multiple arguments. `log` computes a logarithm, and you can specify the base:

```{r}
log(8, base = 2)
```

`8` is a *positional* argument (R knows it's the first one), while `base = 2` is a *named* argument (you're saying explicitly which parameter gets the value 2). Named arguments can go in any order:

```{r}
log(base = 2, x = 8)
```

To learn what arguments a function takes, use `?`:

```{r}
#| eval: false
?log
```

This opens the help page for `log`, which lists its arguments, explains what they do, and gives examples.

`+` is a function too:

```{r}
`+`(3, 5)
```

So is `<-`:

```{r}
`<-`(w, 99)
w
```

Everything that happens in R is a function call. When you write `3 + 5`, R translates it to `` `+`(3, 5) ``: a function called `+`, applied to two arguments. There is no special syntax for arithmetic, no operators baked into the language at a level below functions. In the lambda calculus, the only thing you can do is apply a function to an argument. R took that idea literally.


### Exercises {.unnumbered}

1. Use `round()` to round 3.14159 to two decimal places. (Hint: check `?round` for the argument name.)
2. What does `nchar("functional")` return? What about `nchar("")`?
3. Try `sum(1, 2, 3, 4, 5)`. Now try `sum(1:5)`. Do they give the same result?


## The workspace {#sec-workspace}

Every name you create lives in your *workspace* (also called the global environment\index{environment}). You can see all the names you've defined with `ls()`:

```{r}
#| eval: false
ls()
```

And you can remove a name with `rm()`:

```{r}
#| eval: false
rm(x)
```

After `rm(x)`, the name `x` is gone. The value it pointed to gets cleaned up by garbage collection if nothing else refers to it.

The workspace is simple on the surface: `ls()` shows what's in it, `rm()` removes things. But there is more going on underneath. Functions create their own local environments when they run, and R looks up names by searching through a chain of environments from the inside out. You have been working in one environment so far, the global one. When you start writing functions (@sec-closures-and-scope), the picture gets richer.

::: callout-note
## Saving your workspace
When you quit R, it may ask whether to "save your workspace image." This saves all names and values currently in your workspace to a file, so they reappear next time you open R. In the early days, when computations could take hours, this made sense. Today, most computations are fast enough that you're better off keeping your work in a script file (`.R` or `.qmd`) and rerunning it from scratch. That way, every value in your workspace traces back to a line of code you can point to, and you never wonder where a mysterious leftover variable came from.
:::

Try this:

```{r}
#| eval: false
environment()
```

R returns `<environment: R_GlobalEnv>`. You can pass it to `ls.str()` and get back every binding in it. The container that holds your names is itself a value with a name.
