# Functions are values {#sec-functions-are-values-chapter}

```{r}
#| include: false
source(here::here("_common.R"))
```

In @sec-functions-are-values, you saw that a function is a value\index{first-class function}: you can assign `sqrt` to a new name and call it, the same way you'd assign a number. If functions are values, you can pass them to other functions, store them in lists, and build functions that create new functions.


## Passing functions to functions {#sec-passing-functions}

`sapply()`\index{sapply()} takes a vector and a function, and applies the function to every element:

```{r}
sapply(1:5, sqrt)
```

`sqrt` has no parentheses. You are handing the function itself to `sapply`, which calls it five times, once per element. This is the same distinction from @sec-programs-and-proofs: `fact` is the definition (a rule), while `fact(n)` is the rule applied to an argument.

A loop is still running inside `sapply`, but you never had to write it. You handed over the function and let `sapply` do the iterating.

```{r}
sapply(c(-3, 0, 4, -1, 7), abs)
```

```{r}
sapply(c("hello", "world"), nchar)
```

Any function that takes a single argument works here: `abs` computes absolute values, `nchar` counts characters, and you can pass either to `sapply` with the same syntax.

`sapply` applies your function to each element and then tries to simplify the results into a vector. Every example above returned one value per input, so a vector came back. When the results differ in length or type, `sapply` returns a list instead, and `lapply()`\index{lapply()} always returns a list; @sec-iteration-without-loops covers when each is the right choice.

A function that takes another function as an argument\index{function!as argument} is called a *higher-order function*\index{higher-order function}. `sapply` is the first of many in this book.

::: {.callout-note collapse="true"}
## The lambda calculus connection
In lambda calculus notation (@sec-church), passing a function to another function is just application:

```
sapply = λxs. λf. [f(xs₁), f(xs₂), ..., f(xsₙ)]
sapply([1,2,3,4,5])(sqrt) = [sqrt(1), sqrt(2), sqrt(3), sqrt(4), sqrt(5)]
```

Replace each `f(xsᵢ)` with its value, and the expression reduces to the result vector. This is the same term replacement from @sec-atomic-vectors.
:::


### Exercises {.unnumbered}

1. Use `sapply()` to compute `log()` of the numbers 1 through 10.
2. What happens if you write `sapply(1:5, sqrt())` with parentheses? Try it and read the error.
3. Use `sapply()` and `is.na()` to check which elements of `c(1, NA, 3, NA, 5)` are missing.


## Anonymous functions {#sec-anonymous-functions}

You can pass a named function like `sqrt` to `sapply`. But suppose you need to square every element: there is no built-in `square` function sitting in base R, and creating one just to use it once feels like overkill.

```{r}
sapply(1:5, function(x) x^2)
```

`function(x) x^2` creates a function right where it is needed, with no name and no assignment; it exists only for this one call to `sapply`, and then it vanishes. This is an *anonymous function*\index{anonymous function}, and you already saw a brief example in @sec-functions-are-values.

Since R 4.1 there is a shorter way to write the same thing:

```{r}
sapply(1:5, \(x) x^2)
```

`\(x)` is shorthand for `function(x)`. The backslash is meant to evoke the Greek letter lambda (λ), which is where the whole idea comes from (@sec-church): in lambda calculus this function is written `λx. x²`.

Anonymous functions shine when the operation is short and you only need it once:

```{r}
sapply(c(10, 20, 30), \(x) x / sum(c(10, 20, 30)))
```

```{r}
sapply(c("alice", "bob"), \(name) paste("Hello,", name))
```

When the function gets longer than one line, or when you need it in more than one place, give it a name.

::: callout-tip
## Opinion
If you use a function twice, name it. Anonymous functions belong inside `sapply`, `lapply`, and similar one-off contexts. The moment you catch yourself copying the same `\(x) ...` expression, that function is asking for a name.
:::


### Exercises {.unnumbered}

1. Use `sapply()` with an anonymous function to add 10 to each element of `c(1, 5, 9)`.
2. Rewrite `sapply(1:5, \(x) x^2)` using the older `function(x)` syntax. Verify you get the same result.
3. Write a named function `to_celsius` that converts Fahrenheit to Celsius. Then use `sapply()` to convert `c(32, 72, 100, 212)`.


## Storing functions in lists {#sec-functions-in-lists}

Consider a data frame with mixed column types: some numeric, some character, some logical. You want a one-number summary of each column, but `mean` is meaningless for character data and `length(unique(...))` is meaningless for logicals. The brute-force solution is a chain of `if`/`else if` blocks that checks each column's type and calls the right function. That works for three types and gets ugly at six, and every new type means hunting through the chain for the right insertion point.

The summaries themselves are fine; what tangles is the control flow around them. If functions are values, you can store the summaries in a list, keyed by type, and let a single lookup replace the entire chain.

```{r}
summarizers <- list(
  numeric = \(x) mean(x, na.rm = TRUE),
  character = \(x) length(unique(x)),
  logical = \(x) sum(x, na.rm = TRUE)
)

summarizers$numeric(c(1, 2, 3, NA, 5))
summarizers$character(c("a", "b", "a", "c"))
summarizers$logical(c(TRUE, FALSE, TRUE, TRUE))
```

`summarizers` is a list of three functions, accessed with `$` like any named list element. A list of functions keyed by name is a *dispatch table*: given a key, it returns the function you need. Adding a new type means adding one entry to the list, and the entries that already work never see the change. Package internals, Shiny applications, and testing frameworks all use this pattern to select behavior at runtime.

::: {.callout-note collapse="true"}
## Why dispatch tables feel natural in R
In Church's lambda calculus, functions were values from the start, so a table of them is the ordinary way to hold several alternatives. The `if`/`else` chain is the habit of languages where functions could only be called, never stored.
:::

A smaller table makes the shape easier to see:

```{r}
transforms <- list(
  double = \(x) x * 2,
  triple = \(x) x * 3,
  negate = \(x) -x
)

transforms$double(5)
transforms$triple(5)
transforms$negate(5)
```

The dispatch table stores functions that already exist. But what if you need to manufacture a function that does not exist yet?


### Exercises {.unnumbered}

1. Create a list with two functions: `square` and `cube`. Apply both to the number 4.
2. Write a list of functions where each one converts a temperature from Celsius to a different scale (Fahrenheit, Kelvin). Use them to convert 100 degrees Celsius.


## Functions that return functions {#sec-functions-returning-functions}

A function can return a number, a string, a vector, or a data frame. It can also return a function.

```{r}
make_adder <- function(n) {
  function(x) x + n
}
```

The body of `make_adder` is a single expression, and that expression is a function definition. Call it and see what comes back:

```{r}
add5 <- make_adder(5)
add5(3)
add5(100)
```

```{r}
add10 <- make_adder(10)
add10(3)
```

`add5` and `add10` are both functions, built by `make_adder` with different values of `n`, and each one remembers the value it was created with\index{function!as return value}. A function that manufactures functions this way is a *function factory*. `add5` remembers 5; `add10` remembers 10. How?

When `make_adder(5)` runs, it creates a local environment where `n = 5`. The returned function `function(x) x + n` was defined inside that environment. When you later call `add5(3)`, R finds `n` through lexical scoping (@sec-scope-first-look): it looks in the environment where the function was defined and finds `n = 5`. But that environment was created during a function call that has already returned. The call is gone. So where does the environment live?

It lives with the function. A function that travels with its creation environment is called a *closure*\index{closure}. Every function in R is technically a closure, but the term earns its weight when a function returned by another function captures variables from the enclosing scope. @sec-closures-and-scope shows what else that captured environment can hold: counters, caches, and state that survives between calls.

::: {.callout-note collapse="true"}
## The lambda calculus connection
In lambda calculus notation:

```
make_adder = λn. λx. x + n
make_adder(5) = (λn. λx. x + n)(5) = λx. x + 5
```

Replacing `n` with `5` gives a new function, `λx. x + 5`, which is `add5`. The returned function has one argument already filled in: this is the currying from @sec-arguments-defaults, and each call to a function factory is one beta reduction with the returned function as the reduced term.
:::

```{r}
make_multiplier <- function(factor) {
  function(x) x * factor
}

double <- make_multiplier(2)
triple <- make_multiplier(3)

double(7)
triple(7)
```

`make_multiplier` is another function factory, producing `double` and `triple` from the same template with different values baked in. Function factories appear throughout R, in custom plot themes, parameterized statistical tests, and configured data transformations, and @sec-function-factories covers them in depth. Every function in this chapter so far, passed around, stored in a list, or returned from a factory, has leaned on `+` or `*` to do its work. Are those functions too?


### Exercises {.unnumbered}

1. Write a function `make_power` that takes an exponent `n` and returns a function that raises its argument to the `n`th power. Create `square <- make_power(2)` and `cube <- make_power(3)` and test them.
2. What does `make_adder(0)` return? Is it the identity function?
3. Write a function `make_greeter` that takes a greeting string (like `"Hello"` or `"Bonjour"`) and returns a function that takes a name and produces the greeting. Test: `greet_en <- make_greeter("Hello"); greet_en("Alice")` should return `"Hello, Alice"`.


## Operators are functions {#sec-operators-are-functions}

Try this:

```{r}
`+`(2, 3)
```

`+` is a function. It takes two arguments and returns their sum, and when you write `2 + 3`, R translates it to `` `+`(2, 3) ``. The infix notation is syntactic sugar; underneath, it is a function call.

The same is true for every operator:

```{r}
`*`(4, 5)
`>`(10, 3)
`&`(TRUE, FALSE)
```

It goes further than arithmetic. Indexing with `[` is a function call:

```{r}
x <- c(10, 20, 30)
`[`(x, 2)
```

Accessing a list element with `$` is a function call, and so is assignment with `<-`. Even the curly brace `{` is a function call (it evaluates each expression inside and returns the last one).

John Chambers, the creator of S (@sec-s-language), summarized R's design with two principles:

> "Everything that exists is an object. Everything that happens is a function call."

Numbers, strings, vectors, functions, even `NULL` are all objects. Addition, comparison, indexing, assignment, and control flow are all function calls. Since functions are themselves objects, the two principles fold into each other.

You can even override built-in operators. (This is a demonstration, not a recommendation.)

```{r}
`+` <- function(a, b) a * b
2 + 3
```

```{r}
rm(`+`)
2 + 3
```

After redefining `+` to mean multiplication, `2 + 3` returns 6. `rm()` removes the custom definition and restores the original. The fact that you *can* redefine `+` shows that it really is just a name bound to a function, like any other name, a value you can inspect, replace, and pass around.

::: callout-tip
## Opinion
Never redefine built-in operators in real code. The example above exists to show you what R is made of, not to suggest a workflow. Code that redefines `+` is code that nobody, including you six months later, will be able to read.
:::


### Exercises {.unnumbered}

1. Rewrite `10 - 3` as an explicit function call using backticks.
2. Rewrite `x[1]` as an explicit function call. (Hint: `` `[`(x, 1) ``.)
3. What does `` `{`(1, 2, 3) `` return? Why?
