# Lists {#sec-lists}

```{r}
#| include: false
source(here::here("_common.R"))
```

You have a penguin. It has a species (character), a body mass (double), and a vector of bill and flipper measurements (also doubles). Try cramming all of that into a single atomic vector and R will coerce everything to character, because vectors are strict: one type, no exceptions (@sec-coercion). The species survives, the numbers become strings, and the measurements lose their identity as a group. What you need is a container that keeps each piece intact, preserving its type and its shape, without flattening or coercing anything. R calls that container a list.


## Why lists exist {#sec-why-lists}

Where a vector is homogeneous, a list is heterogeneous: each element can be a number, a string, a vector of arbitrary length, a function, or even another list, and all of them coexist without interference.

```{r}
penguin <- list(
  species = "Adelie",
  mass = 3750,
  measurements = c(39.1, 18.7, 181)
)
penguin
```

`species` is a character scalar, `mass` is a numeric scalar, `measurements` is a numeric vector of length 3. No coercion, no flattening. A vector could not hold these together without reducing them all to a single type; a list simply keeps each element as it was.

Technically, a list is a *recursive vector*, where "recursive" means its elements can themselves be lists, allowing arbitrary nesting. R's documentation calls lists "generic vectors" to distinguish them from atomic vectors (@sec-atomic-vectors), and `typeof()` confirms:

```{r}
typeof(penguin)
```

You already used a list in @sec-functions-in-lists when you stored functions as named elements. Lists are the same structure whether they hold numbers, strings, or functions, but accessing the right element requires care.

The idea of pairing two values into one object has deep roots. In lambda calculus, a pair is encoded as `λa.λb.λf. f a b`: a function that captures two values and waits for a selector. Pass it `λx.λy. x` (select first) and you get `a`; pass `λx.λy. y` (select second) and you get `b`. Lisp's `cons`, `car`, and `cdr` are exactly these operations with names. R's `list()` is the practical descendant of this construction: a container built from the idea that you can bundle values together and retrieve them with selectors (`[[1]]`, `[[2]]`, `$name`).


### Exercises {.unnumbered}

1. Create a list with your name (character), your age (numeric), and your three favourite colours (a character vector). Print it.
2. What does `typeof(list(1, "a", TRUE))` return?
3. What happens if you use `c()` instead of `list()` to combine `1`, `"a"`, and `TRUE`? Why is the result different?


## Creating and accessing lists {#sec-list-access}

### Creating lists

`list()` creates a list, and elements can be named or unnamed:

```{r}
named <- list(a = 1, b = "hello", c = TRUE)
unnamed <- list(1, "hello", TRUE)
```

Named elements are easier to work with because you can refer to them by meaning rather than position; unnamed elements force you to remember which slot holds what.

### The train analogy

Think of a list as a train where each carriage holds its own cargo, and your choice of accessor determines whether you detach the carriage or reach inside it:

- `x[1]` returns the first carriage still attached to the train. The result is a list of length 1.
- `x[[1]]` opens the first carriage and pulls out what's inside. The result is the element itself.
- `x$name` is shorthand for `x[["name"]]`.

```{r}
x <- list(a = 10, b = c(1, 2, 3), c = "hello")
```

`[` returns a sub-list:

```{r}
x[1]
typeof(x[1])
```

Still a list, just a shorter one.

`[[` extracts the element:

```{r}
x[[1]]
typeof(x[[1]])
```

Now it's the number 10, not a list containing 10. This distinction is the single most common source of confusion with lists: `[` keeps the container, `[[` removes it.

`$` works with names:

```{r}
x$b
x[["b"]]
```

Both return the same thing. `$` is convenient for interactive use, but `[[` is necessary when the name is stored in a variable:

```{r}
key <- "b"
x[[key]]
```

```{r}
#| error: true
x$key
```

`x$key` looks for a literal element named `"key"`, not for the value stored in the variable `key`. Whenever the name is computed or passed as an argument, you need `[[`.

You can select multiple elements with `[`:

```{r}
x[c(1, 3)]
x[c("a", "c")]
```

But `[[` only works with a single index, extracting one element at a time.

::: callout-tip
## Opinion
Default to `[[` and `$` for extracting elements. Use `[` only when you need a sub-list. If you find yourself writing `x[1][[1]]`, you wanted `x[[1]]` all along.
:::


### Exercises {.unnumbered}

1. Given `x <- list(a = 10, b = 20, c = 30)`, predict the output of `x[2]`, `x[[2]]`, and `x$b` before running them.
2. What is `typeof(x[1])` versus `typeof(x[[1]])`? Explain the difference.
3. Create a variable `name <- "c"`. Use it to extract the element `"c"` from the list `x`. Which accessor works?


## Nested lists {#sec-nested-lists}

A list element can be another list, which means that a single list can grow into a tree of arbitrary depth:

```{r}
study <- list(
  site = "Palmer Station",
  years = 2007:2009,
  species = list(
    list(name = "Adelie", count = 152),
    list(name = "Gentoo", count = 124),
    list(name = "Chinstrap", count = 68)
  )
)
```

`str()` reveals the shape:

```{r}
str(study)
```

If you only learn one function for working with lists, make it `str()`. It prints the structure compactly, showing types, lengths, and nesting levels at a glance, and it works on any R object, not just lists. Use it whenever you receive something unfamiliar. Every nested list is a tree: the top-level list is the root, each sub-list is a branch, and the atomic values are leaves. The same shape appears in file systems (directories containing files and subdirectories), in HTML (the DOM), and in JSON.

Inspecting is only half the problem. How do you reach into the tree and pull out what you need? Chain `[[` or `$`:

```{r}
study$species[[1]]$name
study[["species"]][[2]][["count"]]
```

Each `[[` or `$` steps one level deeper: `study$species` is a list of three lists, `study$species[[1]]` is the first of those, and `study$species[[1]]$name` is the string `"Adelie"`. JSON from a web API, the output of `lm()`, and configuration files are all nested lists or objects built on them, and the access pattern is always the same: `$` or `[[` to step down one level, repeated as many times as needed.




### Exercises {.unnumbered}

1. Given the `study` list above, extract the count for Chinstrap penguins.
2. Use `str()` on the result of `lm(mpg ~ wt, data = mtcars)`. How many top-level elements does the model object have?
3. Create a nested list representing a book: title, author, and a list of chapters (each with a number and a title). Extract the title of the second chapter.


## Lists as the backbone of R {#sec-lists-backbone}

You might think of lists as a container you reach for occasionally, but in truth they are the foundation that most of R's familiar structures are built on.

A data frame is a list:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
typeof(df)
is.list(df)
```

Each column is one element of that list, and the data frame simply adds a constraint: all columns must have the same length. Underneath, `df$x` works exactly like list access because it *is* list access.

A linear model is a list:

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
typeof(fit)
names(fit)
```

`fit$coefficients`, `fit$residuals`, `fit$fitted.values`: all list extractions. The model object is a list with a class attribute (`"lm"`) that tells R how to print and summarize it, but the data access works the same way as with any other list.

Data frames (@sec-data-frames), model objects, and environments are all lists (or list-like structures) with different rules about what they can contain. The most interesting rule is also the simplest: what happens when every element is a vector of the same length?




### Exercises {.unnumbered}

1. Run `typeof()` on a data frame you create. Then run `is.list()` on it. What do you conclude?
2. Fit a model with `lm(Sepal.Length ~ Petal.Length, data = iris)`. Use `names()` to see its elements, then extract the R-squared from `summary()` of the fit. (Hint: `str(summary(fit))` will help.)
3. What does `length()` return for a data frame with 5 columns and 100 rows? Why?


## Modifying lists {#sec-modifying-lists}

Adding an element is as simple as assigning to a new name:

```{r}
x <- list(a = 1, b = 2)
x$c <- 3
x[["d"]] <- "new"
str(x)
```

Removing an element means setting it to `NULL`:

```{r}
x$b <- NULL
str(x)
```

`b` is gone, not set to `NULL`. This is a common trap: assigning `NULL` to a list element deletes it entirely. If you actually want to store `NULL` as a value, you need `x["e"] <- list(NULL)`:

```{r}
x["e"] <- list(NULL)
str(x)
```

Now `e` exists and its value is `NULL`. The single-bracket assignment with `list(NULL)` is the only way to store an actual `NULL` in a list, an asymmetry worth memorizing because it bites everyone at least once.

Replacing an element is just assignment to an existing name:

```{r}
x$a <- 100
x$a
```

R uses copy-on-modify for lists, the same semantics as for vectors. When you modify a list, R copies only the parts that change, not the entire structure; for practical purposes you can treat assignment as modifying in place, since the copying is an implementation detail that rarely affects your code.


### Exercises {.unnumbered}

1. Create a list with elements `x = 1` and `y = 2`. Add an element `z = 3`, then delete `x`. Print the result.
2. What does `length()` return after you delete an element from a list?
3. Try `x$a <- NULL` on a list where `a` exists. Then try `x["a"] <- list(NULL)`. What is the difference?


## Linked lists {#sec-linked-lists}

Suppose you want to prepend an element to a sequence in constant time, or process a collection by peeling off one element at a time and recursing on the rest. R's built-in lists are arrays of pointers with O(1) random access, which is great for indexing but means prepending requires copying the whole pointer array. A linked list trades random access for efficient prepending: each node holds a value and a pointer to the next node, and the chain ends with `NULL`.

You can build one in R using nested lists:

```{r}
cons <- function(head, tail) list(head = head, tail = tail)
car  <- function(lst) lst$head
cdr  <- function(lst) lst$tail
```

`cons` constructs a node, `car` extracts the first element, and `cdr` extracts the rest. These names come from Lisp (1958), where `car` and `cdr` referred to hardware registers on the IBM 704.

Build a linked list of three elements:

```{r}
ll <- cons(1, cons(2, cons(3, NULL)))
str(ll)
```

To traverse it, recurse until you hit `NULL`:

```{r}
ll_to_vector <- function(lst) {
  if (is.null(lst)) return(c())
  c(car(lst), ll_to_vector(cdr(lst)))
}
ll_to_vector(ll)
```

This structure is a direct translation of Church encoding. In lambda calculus, a pair is `λh.λt.λf. f h t`: a function that captures two values and waits for a selector. `car` passes a selector that returns the first value; `cdr` passes one that returns the second. A linked list is a chain of such pairs, terminated by a special "nil" value. Lisp was built on exactly this encoding, and `cons`/`car`/`cdr` remain the standard names for these operations across functional languages.

So why doesn't R use linked lists for everyday work? R was designed for column-oriented numerical computing, where you need to pass entire columns to C routines like `sum()` or to BLAS, and that requires contiguous memory: a single block of doubles that a C function can walk with a pointer. A linked list scatters its elements across the heap, so you would need to copy them into a contiguous buffer before any vectorized operation could touch them. R's built-in lists (VECSXP) are arrays of pointers, giving O(1) random access and contiguous pointer storage. Linked lists shine at prepending (O(1), just `cons` a new head) and at recursive processing where you always work with the first element and pass the rest along. Lisp used them because Lisp's model is recursive symbolic processing, not numerical arrays; R chose arrays because its model is vectorized computation.


### Exercises {.unnumbered}

1. Using `cons`, `car`, and `cdr` defined above, build the linked list `(10, 20, 30)` and extract the second element without converting to a vector.
2. Write a function `ll_length` that counts the number of nodes in a linked list by recursing through `cdr` until `NULL`.
3. Write a function `ll_map` that takes a linked list and a function, and returns a new linked list with the function applied to each element. Test it by doubling every element of `cons(1, cons(2, cons(3, NULL)))`.


## Looking ahead

Lists hold anything, including other lists. Constrain that freedom (require every element to be a vector of the same length) and the general-purpose container becomes a table.
