# Lists {#sec-lists}

```{r}
#| include: false
source(here::here("_common.R"))
```

You have a penguin. It has a species (character), a body mass (double), and a vector of bill and flipper measurements (also doubles). Try cramming all of that into a single atomic vector and R will coerce everything to character, because vectors are strict: one type, no exceptions (@sec-coercion). The species survives, the numbers become strings, and the measurements lose their identity as a group. What you need is a container that keeps each piece intact, preserving its type and its shape, without flattening or coercing anything. R calls that container a list\index{list}.


## Why lists exist {#sec-why-lists}

Put the three pieces in a list and each one keeps its type:

```{r}
penguin <- list(
  species = "Adelie",
  mass = 3750,
  measurements = c(39.1, 18.7, 181)
)
penguin
```

`species` is still a character scalar, `mass` a numeric scalar, `measurements` a numeric vector of length 3. Nothing was coerced. This is what makes a list heterogeneous\index{heterogeneous}: each element can be a number, a string, a vector of any length, a function, or another list, and none of them interferes with the others.

`typeof()` puts a name on the container:

```{r}
typeof(penguin)
```

R's documentation calls this a *generic vector*, to set it apart from the atomic vectors of @sec-atomic-vectors.

You already used one in @sec-functions-in-lists when you stored functions as named elements. A list is the same structure whether it holds numbers, strings, or functions; getting the right element back out is where the care is needed.


### Exercises {.unnumbered}

1. Create a list with your name (character), your age (numeric), and your three favourite colours (a character vector). Print it.
2. What does `typeof(list(1, "a", TRUE))` return?
3. What happens if you use `c()` instead of `list()` to combine `1`, `"a"`, and `TRUE`? Why is the result different?


## Creating and accessing lists {#sec-list-access}

### Creating lists

`list()` creates a list, and elements can be named or unnamed:

```{r}
named <- list(a = 1, b = "hello", c = TRUE)
unnamed <- list(1, "hello", TRUE)
```

Named elements are easier to work with because you can refer to them by meaning rather than position; unnamed elements force you to remember which slot holds what.

### The train analogy

```{r}
x <- list(a = 10, b = c(1, 2, 3), c = "hello")
```

Ask for the first element with one bracket:

```{r}
x[1]
typeof(x[1])
```

Still a list, just a shorter one. Now with two:

```{r}
x[[1]]
typeof(x[[1]])
```

Now it is the number 10, with no list around it. Think of the list as a train where each carriage holds its own cargo. `x[1]`\index{[ (single bracket)} returns the first carriage still attached to the train, and `x[[1]]`\index{[[ (double bracket)} opens the carriage and pulls out what is inside.

![`x[1]` returns the carriage (still a list); `x[[1]]` pulls out the cargo (the element itself).](../images/list_train.png){#fig-list-train width=100%}

`[` keeps the container and `[[` removes it, and mixing the two up is the most common source of confusion with lists.

`$`\index{\$ (dollar sign)} is shorthand for `[[` with a name:

```{r}
x$b
x[["b"]]
```

Both return the same thing. `$` is convenient for interactive use, but `[[` is necessary when the name is stored in a variable:

```{r}
key <- "b"
x[[key]]
```

```{r}
#| error: true
x$key
```

`x$key` looks for a literal element named `"key"`, not for the value stored in the variable `key`. Whenever the name is computed or passed as an argument, you need `[[`.

You can select multiple elements with `[`:

```{r}
x[c(1, 3)]
x[c("a", "c")]
```

But `[[` only works with a single index, extracting one element at a time.

::: callout-tip
## Opinion
Default to `[[` and `$` for extracting elements. Use `[` only when you need a sub-list. If you find yourself writing `x[1][[1]]`, you wanted `x[[1]]` all along.
:::


### Exercises {.unnumbered}

1. Given `x <- list(a = 10, b = 20, c = 30)`, predict the output of `x[2]`, `x[[2]]`, and `x$b` before running them.
2. What is `typeof(x[1])` versus `typeof(x[[1]])`? Explain the difference.
3. Create a variable `name <- "c"`. Use it to extract the element `"c"` from the list `x`. Which accessor works?


## Nested lists {#sec-nested-lists}

A list element can be another list\index{list!nested}, which means that a single list can grow into a tree of arbitrary depth:

```{r}
study <- list(
  site = "Palmer Station",
  years = 2007:2009,
  species = list(
    list(name = "Adelie", count = 152),
    list(name = "Gentoo", count = 124),
    list(name = "Chinstrap", count = 68)
  )
)
```

`str()`\index{str()} reveals the shape:

```{r}
str(study)
```

`str()` is the single most useful function for lists. It works on any R object, but with lists it shows you the nesting, the types, and the lengths in one compact summary. Get in the habit of calling it on anything you did not build yourself.

This is why R's documentation also calls lists *recursive vectors*\index{recursive data structure}: an element can itself be a list, so one list can grow into a tree of any depth.

Inspecting is only half the problem. How do you reach into the tree and pull out what you need? Chain `[[` or `$`:

```{r}
study$species[[1]]$name
study[["species"]][[2]][["count"]]
```

Each `[[` or `$` steps one level deeper: `study$species` is a list of three lists, `study$species[[1]]` is the first of those, and `study$species[[1]]$name` is the string `"Adelie"`. Sub-lists are branches and atomic values are leaves, which is also the shape of a file system, of a JSON\index{JSON} document from a web API, and of the object `lm()` returns. All of them read into R as nested lists because all of them are trees.




### Exercises {.unnumbered}

1. Given the `study` list above, extract the count for Chinstrap penguins.
2. Use `str()` on the result of `lm(mpg ~ wt, data = mtcars)`. How many top-level elements does the model object have?
3. Create a nested list representing a book: title, author, and a list of chapters (each with a number and a title). Extract the title of the second chapter.


## Lists as the backbone of R {#sec-lists-backbone}

A data frame is a list:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
typeof(df)
is.list(df)
```

Each column is one element of that list, and the data frame simply adds a constraint: all columns must have the same length. Underneath, `df$x` works exactly like list access because it *is* list access.

A linear model is a list:

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
typeof(fit)
names(fit)
```

`fit$coefficients`, `fit$residuals`, and `fit$fitted.values` are ordinary `$` extractions on a list. The `"lm"` class attribute changes how R prints and summarizes the object, not how you get data out of it.

Data frames (@sec-data-frames), model objects, and environments are all lists (or list-like structures) with different rules about what they can contain. The most interesting rule is also the simplest: what happens when every element is a vector of the same length?




### Exercises {.unnumbered}

1. Run `typeof()` on a data frame you create. Then run `is.list()` on it. What do you conclude?
2. Fit a model with `lm(Sepal.Length ~ Petal.Length, data = iris)`. Use `names()` to see its elements, then extract the R-squared from `summary()` of the fit. (Hint: `str(summary(fit))` will help.)
3. What does `length()` return for a data frame with 5 columns and 100 rows? Why?


## Modifying lists {#sec-modifying-lists}

Adding an element is as simple as assigning to a new name:

```{r}
x <- list(a = 1, b = 2)
x$c <- 3
x[["d"]] <- "new"
str(x)
```

Removing an element means setting it to `NULL`:

```{r}
x$b <- NULL
str(x)
```

`b` is gone, not set to `NULL`. This is a common trap: assigning `NULL` to a list element deletes it entirely. If you actually want to store `NULL` as a value, you need `x["e"] <- list(NULL)`:

```{r}
x["e"] <- list(NULL)
str(x)
```

Now `e` exists and its value is `NULL`. The single-bracket assignment with `list(NULL)` is the only way to store an actual `NULL` in a list, an asymmetry worth memorizing because it bites everyone at least once.

Replacing an element is just assignment to an existing name:

```{r}
x$a <- 100
x$a
```

R uses copy-on-modify for lists, the same semantics as for vectors (@sec-space-complexity): it copies only the parts that change, not the entire structure, so for practical purposes you can treat assignment as modifying in place.


### Exercises {.unnumbered}

1. Create a list with elements `x = 1` and `y = 2`. Add an element `z = 3`, then delete `x`. Print the result.
2. What does `length()` return after you delete an element from a list?
3. Try `x$a <- NULL` on a list where `a` exists. Then try `x["a"] <- list(NULL)`. What is the difference?


## Linked lists {#sec-linked-lists}

Three one-line functions are enough to build a data structure R does not ship with:

```{r}
cons <- function(head, tail) list(head = head, tail = tail)
car  <- function(lst) lst$head
cdr  <- function(lst) lst$tail
```

`cons` builds a node holding a value and whatever comes after it, `car` returns the value, and `cdr` returns the rest. Chain three nodes:

```{r}
ll <- cons(1, cons(2, cons(3, NULL)))
str(ll)
```

Each node is a two-element list whose `tail` is the next node, and the last `tail` is `NULL`. This is a linked list\index{linked list}. To walk it, recurse until you hit `NULL`:

```{r}
ll_to_vector <- function(lst) {
  if (is.null(lst)) return(c())
  c(car(lst), ll_to_vector(cdr(lst)))
}
ll_to_vector(ll)
```

Adding a node at the front costs one `cons` call however long the chain is, while reaching element *k* means following *k* `tail` pointers. An R vector is the reverse: element *k* is one address computation away, and putting a value in front copies everything. R does not use linked lists for everyday work because `sum()`, and the BLAS routines behind matrix algebra, expect one contiguous block of doubles that a C loop can walk with a pointer, and a linked list scatters its nodes across memory. R's own lists are contiguous arrays of pointers too, which is why `x[[k]]` on a list is O(1). Lisp made the cons cell its primary structure instead, which is where the names come from.

::: {.callout-note collapse="true"}
## From Church encoding to cons cells

The names `car` and `cdr` come from Lisp, where in 1958 they abbreviated "contents of the address register" and "contents of the decrement register", the two halves of a machine word on the IBM 704. The structure itself is older. In lambda calculus a pair is `λh.λt.λf. f h t`, a function that captures two values and waits for a selector; `car` passes a selector that returns the first value and `cdr` one that returns the second. A linked list is a chain of such pairs ending in a nil value, and `cons`, `car`, and `cdr` remain the standard names for the three operations across functional languages.

R and Lisp descend from the same calculus, but the engineering diverged early. Lisp built up answers one `cons` at a time, peeling the head off a list and recurring on the tail. R committed to contiguous arrays that the C runtime walks in one pass, and that choice propagated: R has `sapply` where Lisp has `mapcar`, and `[` extracts by position where Lisp peels from the front. The linked-list code above shows what the other branch of the family looks like from the inside.
:::


### Exercises {.unnumbered}

1. Using `cons`, `car`, and `cdr` defined above, build the linked list `(10, 20, 30)` and extract the second element without converting to a vector.
2. Write a function `ll_length` that counts the number of nodes in a linked list by recursing through `cdr` until `NULL`.
3. Write a function `ll_map` that takes a linked list and a function, and returns a new linked list with the function applied to each element. Test it by doubling every element of `cons(1, cons(2, cons(3, NULL)))`.


## Looking ahead

Lists hold anything, including other lists. Constrain that freedom (require every element to be a vector of the same length) and the general-purpose container becomes a table.
