# Strings, factors, and dates {#sec-strings-factors-dates}

```{r}
#| include: false
source(here::here("_common.R"))
```

You have a column of dates in three different formats, a species column that plots in the wrong order, and a free-text field full of trailing whitespace and inconsistent capitalization. Strings, factors, and dates each get a section here and a tidyverse package of their own. The three sections are independent; skip to whichever one is causing you trouble right now.

We will use the `palmerpenguins` dataset throughout. Load it and the three packages now:

```{r}
library(palmerpenguins)
library(stringr)
library(forcats)
library(lubridate)
```


## Character vectors {#sec-character-vectors}

```{r}
x <- "hello"
typeof(x)
length(x)
```

A string\index{string}\index{character} in R is a character vector of length 1. There is no separate string type; `"hello"` is `character(1)`, the same kind of atomic vector you met in @sec-atomic-vectors. Double quotes and single quotes both work. Pick one and stick with it.

::: callout-tip
## Opinion
Use double quotes. R's own style guide does, and so does most code you will encounter in the wild. Single quotes earn their keep inside double-quoted strings: `"it's easy"`.
:::

R uses UTF-8\index{UTF-8}\index{encoding} as its modern string encoding. If accented characters come out garbled, the file is probably Latin-1; `stringr::str_conv()` or `readr::locale(encoding = "latin1")` handles that.

::: {.callout-note collapse="true"}
## Where UTF-8 comes from

UTF-8 was designed by Ken Thompson and Rob Pike on a placemat in a New Jersey diner in September 1992. It is backward-compatible with ASCII, self-synchronizing, and variable-width (one to four bytes per character).
:::

Special characters use backslash escapes: `\n` (newline), `\t` (tab), `\\` (literal backslash). Compare `print()` and `cat()`:

```{r}
print("line one\nline two")
cat("line one\nline two")
```

`print()` shows the escape sequence as literal text; `cat()` renders it. A small difference, until you spend ten minutes wondering why your output has `\n` in it instead of a line break.

Base R provides a handful of string tools:

```{r}
nchar("penguin")
paste("Gentoo", "penguin")
paste0("Gentoo", "penguin")
sprintf("The %s weighs %d grams", "Gentoo", 5200)
```

`nchar()` counts characters, `paste()`\index{paste()} joins with a space, `paste0()` joins without, and `sprintf()` does formatted substitution borrowing syntax from C. These work, but they are inconsistent in argument order and `NA` handling, and inconsistency compounds fast once you start chaining operations. What would a consistent interface look like?


### Exercises {.unnumbered}

1. What does `nchar(NA)` return? What about `nchar("")`?
2. Use `paste()` to combine `"Species"`, `":"`, and `"Adelie"` into a single string. Then do the same with `paste0()`. What is different?
3. Use `sprintf()` to produce the string `"Island: Biscoe, n = 168"`.


## stringr: consistent string operations {#sec-stringr}

```{r}
str_length("penguin")
str_sub("penguin", 1, 4)
str_c("Gentoo", "penguin", sep = " ")
```

`str_length()` counts characters like `nchar()`, `str_sub()` extracts by position, and `str_c()` combines strings. Every function in stringr\index{stringr (package)} follows the same rule: the name starts with `str_`, the string is the first argument, and the pattern, when there is one, comes second. Once you know the convention you can guess a function's name before looking it up. Where `paste()` turns a missing value into the text `"NA"`, `str_c()` keeps it missing:

```{r}
paste("hello", NA)
str_c("hello", NA)
```

Case conversion and whitespace cleaning:

```{r}
str_to_upper("gentoo")
str_to_title("gentoo penguin")
str_trim("  messy data  ")
str_squish("  too   many   spaces  ")
```

`str_detect()` answers the question `grepl()`\index{grep()} answers in base R, whether each string contains the pattern:

```{r}
species <- penguins$species
str_detect(species, "Gentoo")[1:10]
```

Extraction and replacement:

```{r}
islands <- c("Biscoe Island", "Dream Island", "Torgersen Island")
str_extract(islands, "^[A-Za-z]+")
str_replace(islands, "Island", "Isl.")
```

`str_replace()` changes the first match in each string and `str_replace_all()` every match, the way `sub()` and `gsub()`\index{gsub()} do in base R. Splitting:

```{r}
str_split("one-two-three", "-")
```

`str_split()` returns a list because each input string could split into a different number of pieces. A list is the container for output of uneven length (@sec-lists), which is also why `strsplit()` returns one in base R.

On the penguins data:

```{r}
str_to_upper(levels(penguins$island))
sum(str_detect(penguins$species, "Adelie"))
```



### Exercises {.unnumbered}

1. How many penguins have species names containing the letter "e"? Use `str_detect()`.
2. Use `str_sub()` to extract the first three letters of each island name in `penguins$island`.
3. Use `str_replace_all()` to replace all spaces in `"Gentoo penguin species"` with underscores.


## Regex essentials {#sec-regex-essentials}

`str_detect()` has so far matched literal text. Ask it for every string that starts with a capital letter and it needs a different kind of pattern:

```{r}
str_detect(c("Hello", "world", "R"), "^[A-Z]")
```

`^` anchors the match to the start of the string and `[A-Z]` matches one capital letter. A pattern like this is a regular expression\index{regular expression}\index{regex}, a small language for describing text\index{pattern matching}, and the same patterns work in Python, JavaScript, and `grep` on the command line. `str_view()` shows what a pattern matches, which makes it the tool to build one with:

```{r}
fruits <- c("apple", "banana", "cherry", "date", "elderberry")
str_view(fruits, "[aeiou]")
```

Two more that come up constantly:

```{r}
# Strings that end in a digit
str_detect(c("room101", "lobby", "floor3"), "\\d$")

# Extract numbers from text
str_extract("penguin weighs 5200 grams", "\\d+")
```

`\\d` is any digit, `$` anchors to the end of the string, and `+` means one or more of whatever precedes it. The doubled backslash exists because R strings process backslashes first: to get `\d` to the regex engine, you write `"\\d"` in R. The building blocks used so far, and the handful of others you will need:

| Pattern | Matches |
|---------|---------|
| `.` | Any single character |
| `^` | Start of string |
| `$` | End of string |
| `[abc]` | Any of a, b, or c |
| `[0-9]` | Any digit |
| `+` | One or more of the preceding |
| `*` | Zero or more of the preceding |
| `?` | Zero or one of the preceding |
| `(A|B)` | A or B (alternation) |
| `\\d` | Digit (same as `[0-9]`) |
| `\\s` | Whitespace |
| `\\w` | Word character (letter, digit, underscore) |

`?regex` has the full reference.

::: {.callout-note collapse="true"}
## Where regular expressions come from

Regular expressions come from formal language theory, where Stephen Kleene\index{Kleene, Stephen} defined regular languages in 1956 using concatenation, alternation, and closure (repetition). Ken Thompson implemented them in the QED editor in 1968 and `grep` in 1973; the `*` quantifier is still called the *Kleene star*, after Kleene.
:::

::: callout-tip
## Opinion
You do not need to memorize regex. Know that it exists, know the basics from the table above, and keep the stringr cheatsheet within arm's reach. That is enough.
:::



### Exercises {.unnumbered}

1. Write a regex that matches strings starting with "G" and ending with "o". Test it on `c("Gentoo", "Galileo", "Go", "Gusto", "Goo")`.
2. Use `str_extract_all()` to pull all words (sequences of `\\w+`) from `"The quick brown fox"`.
3. Use `str_detect()` and a regex to find which island names in `penguins$island` contain two consecutive vowels.


## Why factors exist {#sec-why-factors}

Strings handle free text. Species, island, and treatment group are a different kind of column: a fixed set of values, and an order among them that matters for plots and models. Plot penguin species as a bar chart and the bars come out alphabetically, Adelie, Chinstrap, Gentoo. A character vector carries no ordering, so there is no way to tell the plot to put Gentoo first. R's answer is to store such a column as integers with labels:

```{r}
x <- factor(c("male", "female", "female"))
x
typeof(x)
unclass(x)
```

`typeof()` returns `"integer"`, and `unclass()` strips the shell to show what is stored: `2, 1, 1` with a lookup table, `1 = "female"` and `2 = "male"`, alphabetical by default. This is a factor\index{factor}.

The lookup table saves memory, and it does something more useful than that. Ten thousand observations of three species are ten thousand integers and a three-entry table rather than ten thousand copies of the string `"Adelie"`. The declared levels are also a contract: once `factor(species, levels = c("Adelie", "Chinstrap", "Gentoo"))` has run, those three values can appear in the column and nothing else, so a typo such as `"adelie"` fails at assignment instead of turning into a fourth species inside a model. The same contract lets `lm()` build a dummy variable for every level, including one absent from the subset you happen to be fitting.

::: {.callout-note collapse="true"}
## Sum types

In type theory a factor is a *sum type*: the structure of a logical value (@sec-logical-vectors), generalized from two variants (`Bool = TRUE | FALSE`) to *n* (`Species = Adelie | Chinstrap | Gentoo`). A data frame combines fields with AND, the product type from the previous chapter; a factor chooses one variant with OR.
:::

The old `data.frame()` default converted strings to factors (@sec-creating-data-frames), which is why so much older code carries the `stringsAsFactors = FALSE` incantation.

Factors earn their place wherever the set of values or their order matters: the order of bars, legends, and facets in a plot; the levels that `lm()` and `glm()` turn into coefficients; months, Likert scales, treatment groups. For filtering and counting, a character vector is fine. The moment a plot needs its bars in an order that tells a story, you will reach for factors, and then the question is how to rearrange the levels without pain.


### Exercises {.unnumbered}

1. Create a factor from `c("low", "medium", "high", "low", "high")`. What are the levels? In what order?
2. Use `unclass()` to see the integer codes. Which integer corresponds to "low"?
3. What happens if you try to assign a value that is not in the levels? Try `x[1] <- "extreme"` on your factor.


## forcats: taming factors {#sec-forcats}

Base R's `factor()` lets you set levels manually:

```{r}
sizes <- factor(c("small", "medium", "large"), levels = c("small", "medium", "large"))
sizes
```

The `levels`\index{levels} argument controls the allowed values and their order. Without it, R defaults to alphabetical, which is why "high" comes before "low" and plots look wrong. But manually specifying levels for every factor in every analysis gets tedious fast, and tedium breeds errors.

forcats\index{forcats (package)} provides cleaner tools, all starting with `fct_`:

```{r}
# Reorder levels manually
fct_relevel(sizes, "large", "medium", "small")
```

```{r}
# Order levels by frequency in the data
fct_infreq(penguins$species) |> table()
```

```{r}
# Collapse rare levels into "Other"
fct_lump_n(penguins$species, n = 2) |> table()
```

```{r}
# Rename levels
fct_recode(penguins$species, AP = "Adelie", GP = "Gentoo", CP = "Chinstrap") |> head()
```

The most useful function is `fct_reorder()`, which reorders levels by a summary of another variable:

```{r}
#| fig.width: 5
#| fig.asp: 0.6
library(ggplot2)

penguins_clean <- penguins[!is.na(penguins$body_mass_g), ]

ggplot(penguins_clean, aes(x = fct_reorder(species, body_mass_g, median), y = body_mass_g)) +
  geom_boxplot() +
  labs(x = "Species (ordered by median body mass)", y = "Body mass (g)")
```

Without `fct_reorder()`, species appear alphabetically: Adelie, Chinstrap, Gentoo. With it, they appear in order of median body mass, and the axis now carries information about the penguins instead of about the alphabet.


### Exercises {.unnumbered}

1. Use `fct_infreq()` on `penguins$island` to see which island has the most observations.
2. Use `fct_lump_n()` with `n = 1` on `penguins$species`. What happens?
3. Create a bar chart of `penguins$species` with bars ordered by frequency (hint: `fct_infreq()` inside `aes()`).


## Dates and times {#sec-dates-times}

Add one month to January 31. February 28? February 31, which does not exist? An error? Reasonable tools disagree, and the disagreement is a hint of what date arithmetic has to handle: months of different lengths, leap years, time zones, and daylight-saving transitions. Start with what R stores:

```{r}
today <- Sys.Date()
today
typeof(today)
unclass(today)
```

A `Date`\index{Date} is a double counting days since 1970-01-01. The epoch comes from Unix, which measured time in seconds from that date, and R, Python, JavaScript, and most databases inherit it. For a date with a time of day, R has `POSIXct`\index{POSIXct}, seconds since the same epoch and the class to use in data frames, and `POSIXlt`, the same instant stored as a named list of components, which you will rarely need.

::: {.callout-note collapse="true"}
## The Year 2038 problem

32-bit signed integers can count seconds from 1970 until January 19, 2038.
:::

Date arithmetic works as you would expect:

```{r}
as.Date("2026-03-07") - as.Date("2026-01-01")
```

R returns a `difftime` object. Subtraction works, and so does addition with scalars:

```{r}
as.Date("2026-01-01") + 30
```

The base R parsing function is `as.Date()`, which expects ISO 8601 format (`"YYYY-MM-DD"`) by default:

```{r}
as.Date("2026-03-07")
as.Date("07/03/2026", format = "%d/%m/%Y")
```

The `format` argument uses `%Y` (4-digit year), `%m` (month), `%d` (day), and similar codes. These are hard to remember and easy to get wrong, especially when your data mixes formats. Is `"03/07/2026"` March 7th or July 3rd? The format string decides, and if you pick wrong, R will not complain.



### Exercises {.unnumbered}

1. What day number (since 1970-01-01) is today? Use `unclass(Sys.Date())`.
2. What date is 1000 days from today? Use `Sys.Date() + 1000`.
3. How many days are between `"2024-02-28"` and `"2024-03-01"`? (2024 is a leap year.)


## lubridate: dates for humans {#sec-lubridate}

Those `%Y/%m/%d` format codes? lubridate\index{lubridate (package)} lets you forget them. Its parsing functions are named after the order of components, so the function name *is* the format:

```{r}
ymd("2026-03-07")
dmy("07/03/2026")
mdy("03-07-2026")
```

All three produce the same date. No format strings to memorize, no ambiguity about which code means what.

For date-times:

```{r}
ymd_hms("2026-03-07 14:30:00")
```

Extracting components:

```{r}
d <- ymd("2026-03-07")
year(d)
month(d)
day(d)
wday(d, label = TRUE)
```

Date arithmetic with human-readable units:

```{r}
d + days(30)
d + months(1)
d + years(1)
```

Now back to January 31:

```{r}
ymd("2026-01-31") + months(1)
ymd("2026-01-31") %m+% months(1)
```

`+ months(1)` lands on a day that does not exist and returns `NA` rather than picking a date for you. `%m+%` is the operator that rolls back to the last valid day, February 28, or 29 in a leap year. Both are defensible answers to the question that opened the previous section; lubridate makes you choose.

The `days(30)` above is a period, a count of calendar days. lubridate also has durations, exact counts of seconds:

```{r}
days(1)
ddays(1)
```

The two differ across a daylight-saving transition, where a calendar day has 23 or 25 hours:

```{r}
x <- ymd_hms("2026-03-28 12:00:00", tz = "Europe/Vienna")
x + days(1)
x + ddays(1)
```

The period lands at noon the next day; the duration lands at one o'clock, 86,400 seconds later. For most work, periods (`days()`, `months()`, `years()`) are what you want, and durations (`ddays()`, `dhours()`) are for physical time, the kind a stopwatch measures. A third kind of span, the interval, is anchored at a start and an end:

```{r}
# How old is R? (first public release: 1993-08-01)
interval(ymd("1993-08-01"), Sys.Date()) %/% years(1)
```

::: callout-tip
## Opinion
Parse with lubridate, extract with lubridate, do arithmetic with lubridate. Reach for base R date functions only when you need zero dependencies.
:::


### Exercises {.unnumbered}

1. Parse the following dates: `"15-Jan-2024"`, `"2024/06/30"`, `"December 25, 2023"`. Which lubridate function does each need?
2. What day of the week were you born? Use `ymd()` and `wday(label = TRUE)`.
3. Compute the number of days between `"2020-03-01"` and `"2026-03-01"`. Then compute the number of months using `interval()` and `%/%`.


## Summary {#sec-strings-factors-dates-summary}

Each of these data types has a matching tidyverse package:

| Data type | Base R | Tidyverse package | Prefix |
|-----------|--------|--------------------|--------|
| Text | `paste()`, `grep()`, `sub()` | stringr | `str_` |
| Categories | `factor()`, `levels()` | forcats | `fct_` |
| Dates | `as.Date()`, `Sys.Date()` | lubridate | `ymd()`, `year()`, ... |

The tidyverse packages give these operations a consistent interface and naming scheme. In some cases, such as `str_detect()` beside `grepl()`, the difference is cosmetic; in others, such as the `NA` handling of `str_c()` or the choice between `+` and `%m+%`, the package makes a decision that base R leaves to you.

Data frames need data, so @sec-reading-writing-data deals with getting files off disk and into R. After that the pieces combine: `str_detect()` inside `filter()`, `fct_reorder()` inside `ggplot()`, date arithmetic inside `mutate()`.
