# Data transformation {#sec-data-transformation}

```{r}
#| include: false
source(here::here("_common.R"))
```

Are Gentoo penguins heavier than Adelie penguins? The `penguins` data has the answer, but getting there means filtering rows, picking columns, computing a mean, and grouping by species. Every data analysis you will ever write is some arrangement of those same moves, and the question that shapes this chapter is: what is the smallest set of operations that covers all of them?


## The idea: verbs for data {#sec-verbs-for-data}

```{r}
library(palmerpenguins)
library(dplyr)
```

The `penguins` data frame (loaded by `palmerpenguins`) has 344 rows and 8 columns: species, island, bill measurements, flipper length, body mass, sex, and year. Every example in this chapter uses it.

```{r}
penguins
```

To answer the opening question you keep the rows that have a body mass, split them by species, and average. dplyr\index{dplyr (package)} gives each of those moves its own function, and every one of them takes a data frame as its first argument and returns a data frame. `filter()`\index{filter()} keeps rows that match a condition, `select()`\index{select()} picks columns by name, `mutate()`\index{mutate()} adds columns computed from existing ones, `summarise()`\index{summarise()} collapses rows into a summary, and `arrange()`\index{arrange()} reorders them. Because each verb hands back the same kind of object it received, you can chain them (via pipes, @sec-pipes-and-composition), and each one does exactly one thing.

::: {.callout-note collapse="true"}
## Codd's algebra

These verbs did not come from nowhere. The 1970 paper that introduced the relational model (@sec-list-of-vectors) also defined a *relational algebra*\index{relational algebra}: a small set of operations on tables where every operation takes a table in and returns a table out, among them restriction (keep rows) and projection (keep columns). IBM resisted at first, because it had a competing database product called IMS. The composability was too useful, and by the 1980s relational databases had won.

dplyr is Codd's algebra with friendlier names. The same ideas appear in SQL, pandas, and data.table because the operations are the same; only the syntax differs. `arrange()` is the one exception: ordering has no relational equivalent, but you need it for display.
:::


## `filter()`: keep rows {#sec-filter}

You have 344 penguins. You only care about the Adelie ones. How do you throw the rest away?

```{r}
filter(penguins, species == "Adelie")
```

The logical expression works the same way as in @sec-comparison-operators. `species` here refers to the column, not to anything in your global environment: dplyr evaluates the expression inside the data frame.

Multiple conditions, separated by commas, are combined with AND:

```{r}
#| results: hide
filter(penguins, species == "Adelie", body_mass_g > 4000)
```

This keeps only Adelie penguins heavier than 4000 grams. You can also write `&` explicitly, but the comma is cleaner.

For OR, use `|`:

```{r}
#| results: hide
filter(penguins, species == "Adelie" | species == "Chinstrap")
```

When testing membership in a set, `%in%` is clearer than chaining `|`:

```{r}
#| results: hide
filter(penguins, species %in% c("Adelie", "Chinstrap"))
```

One detail worth knowing: `filter()` drops rows where the condition evaluates to `NA`. If `body_mass_g` is `NA` for a row, then `body_mass_g > 4000` is `NA`, and `filter()` excludes it. This is usually what you want, but if you need to keep `NA` rows, test for them explicitly: `filter(penguins, is.na(body_mass_g))`.

`filter()` is the logical subsetting from @sec-logical-subsetting-df, `penguins[penguins$species == "Adelie", ]`, with the intent in the function name and a result the next verb can take as input. But keeping rows is only half the story; what about columns you don't need?


### Exercises {.unnumbered}

1. Filter `penguins` to keep only female penguins from Biscoe island.
2. Filter `penguins` to keep rows where `flipper_length_mm` is greater than 200 or `body_mass_g` is less than 3000.
3. How many rows have missing values in `sex`? Use `filter()` and `nrow()`.


## `select()`: pick columns {#sec-select}

Eight columns is manageable. Eighty is not. When you only need species, bill length, and body mass, carrying around five extra columns clutters every printout, slows every join, and obscures what matters. `select()` strips a data frame down to the columns you name.

```{r}
select(penguins, species, bill_length_mm, body_mass_g)
```

The rest are gone.

Drop columns by prefixing with `-`:

```{r}
#| results: hide
select(penguins, -year)
```

For more flexible selection, dplyr provides helper functions. `starts_with()`, `ends_with()`, and `contains()` match column names by pattern:

```{r}
#| results: hide
select(penguins, starts_with("bill"))
```

```{r}
#| results: hide
select(penguins, ends_with("mm"))
```

`where()` selects columns based on their content, not their name. `where(is.numeric)` keeps every numeric column:

```{r}
#| results: hide
select(penguins, species, where(is.numeric))
```

You can rename columns inside `select()`:

```{r}
select(penguins, species, mass = body_mass_g)
```

This selects `species` and `body_mass_g`, renaming the latter to `mass`, and every other column is gone. If you want the new name and the other six columns, use `rename()`:

```{r}
#| results: hide
rename(penguins, mass = body_mass_g)
```

To reorder columns (move `species` to the front, keep everything else), use `everything()`:

```{r}
#| results: hide
select(penguins, island, species, everything())
```


### Exercises {.unnumbered}

1. Select all columns from `penguins` that contain the string `"length"`.
2. Select `species`, `island`, and all numeric columns. (Hint: combine names and `where()`.)
3. What is the difference between `select(penguins, mass = body_mass_g)` and `rename(penguins, mass = body_mass_g)`? Try both.


## `mutate()`: compute new columns {#sec-mutate}

Body mass is recorded in grams, but your collaborator's model expects kilograms.

```{r}
mutate(penguins, mass_kg = body_mass_g / 1000)
```

You can reference a column you just created in the same `mutate()` call:

```{r}
#| results: hide
mutate(
  penguins,
  mass_kg = body_mass_g / 1000,
  mass_lb = mass_kg * 2.205
)
```

`mass_lb` uses `mass_kg`, which was defined one line above. In a plain function call the second argument cannot see the first; `mutate()` evaluates its arguments in order, top to bottom, so each expression can use every column defined above it. Reverse the two lines and `mass_lb` fails, because `mass_kg` does not exist yet.

Any function that works on vectors works inside `mutate()`: `log()`, `sqrt()`, `cumsum()`, `lag()`, `lead()`, `percent_rank()`. The conditional functions from @sec-vectorized-conditionals work here too:

```{r}
#| results: hide
mutate(
  penguins,
  size = if_else(body_mass_g > 4000, "large", "small")
)
```

For more than two categories, `case_when()` is the tool:

```{r}
#| results: hide
mutate(
  penguins,
  size = case_when(
    body_mass_g > 5000 ~ "large",
    body_mass_g > 3500 ~ "medium",
    .default = "small"
  )
)
```

If you want to keep only the new columns and drop everything else, use `.keep = "none"`:

```{r}
mutate(
  penguins,
  mass_kg = body_mass_g / 1000,
  .keep = "none"
)
```


### Exercises {.unnumbered}

1. Add a column `bill_ratio` that divides `bill_length_mm` by `bill_depth_mm`.
2. Use `mutate()` and `case_when()` to create a column `island_code` that maps `"Biscoe"` to `"B"`, `"Dream"` to `"D"`, and `"Torgersen"` to `"T"`.
3. What happens if you use a column name that already exists in `mutate()`? Try `mutate(penguins, species = toupper(species))`.


## `summarise()` and `group_by()`: aggregate {#sec-summarise-group-by}

What is the average body mass of a penguin in this dataset?

```{r}
summarise(penguins, mean_mass = mean(body_mass_g, na.rm = TRUE))
```

One row, one number. The opening question asked for the mean per species, and `summarise()` on its own has no idea what a species is.

```{r}
penguins_grouped <- group_by(penguins, species)
summarise(penguins_grouped, mean_mass = mean(body_mass_g, na.rm = TRUE))
```

The `summarise()` call is unchanged and now returns three rows, one per species. `group_by()`\index{group\_by()} did not touch the data; it attached metadata that tells subsequent verbs to operate per group, and `summarise()` ran `mean()` once inside each\index{split-apply-combine}. Print `penguins_grouped` and the only visible difference is a `Groups:` line in the header.

::: {.callout-note collapse="true"}
## Grouping in other languages

SQL spells this `GROUP BY`; Spark uses explicit partitioning; in functional programming the same idea appears as `foldByKey`, where a reduce operation runs per partition rather than globally. The underlying pattern, an aggregation function paired with a partitioning structure, is the same across all of them.
:::

You can group by multiple variables:

```{r}
summarise(
  group_by(penguins, species, island),
  mean_mass = mean(body_mass_g, na.rm = TRUE),
  count = n()
)
```

`summarise()` prints a message this time: it removed the last grouping variable, `island`, and the result is still grouped by `species`. That leftover grouping is easy to forget about, and the Opinion at the end of this section says what to do with it.

`n()`\index{n()} counts the number of rows in each group; `n_distinct()` counts unique values:

```{r}
summarise(
  group_by(penguins, species),
  n_islands = n_distinct(island)
)
```

The shortcut `count()`\index{count()} does `group_by()` plus `summarise(n = n())` in one step:

```{r}
count(penguins, species)
```

::: callout-tip
## Opinion
After `summarise()`, the last grouping variable is dropped and the result may still be grouped by the others. A `mutate()` you expect to operate on the full data frame then operates per group instead. Add `.groups = "drop"` to `summarise()`, or call `ungroup()` afterward.
:::


### Exercises {.unnumbered}

1. Compute the median `flipper_length_mm` for each species.
2. Use `group_by()` and `summarise()` to find the maximum `bill_length_mm` per species and island combination.
3. Use `count()` to find how many penguins were observed in each year.



## `arrange()`: order rows {#sec-arrange}

`arrange()` sorts rows. The default is ascending:

```{r}
arrange(penguins, body_mass_g)
```

Wrap a column in `desc()` for descending order:

```{r}
#| results: hide
arrange(penguins, desc(body_mass_g))
```

Multiple columns break ties; sort by species alphabetically, then by mass descending within each species:

```{r}
#| results: hide
arrange(penguins, species, desc(body_mass_g))
```

Unlike the other verbs, `arrange()` ignores grouping by default. If you want to sort within groups, set `.by_group = TRUE`.

`arrange()` is pure display. No information is added or removed, only the order of rows changes. But a well-sorted table can reveal patterns that a jumbled one hides, and sometimes the five core verbs are not quite enough to expose them.


## More verbs {#sec-more-verbs}

`distinct()` returns unique rows:

```{r}
distinct(penguins, species, island)
```

Five combinations of species and island exist in the data.

`slice_head()`\index{slice()} and `slice_tail()` keep the first or last `n` rows:

```{r}
slice_head(penguins, n = 3)
```

`slice_max()` and `slice_min()` keep the rows with the largest or smallest values:

```{r}
slice_max(penguins, body_mass_g, n = 3)
```

`pull()` extracts a single column as a vector, the pipe-friendly equivalent of `$`:

```{r}
pull(penguins, species) |> head()
```

`relocate()` moves columns without dropping any. Move `sex` and `year` to the front:

```{r}
#| results: hide
relocate(penguins, sex, year)
```

Finally, `across()`\index{across()} applies a function to multiple columns at once. Summarise all numeric columns by their mean:

```{r}
summarise(penguins, across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

`across()` works inside `mutate()` too. Round every numeric column to one decimal place:

```{r}
#| results: hide
mutate(penguins, across(where(is.numeric), \(x) round(x, 1)))
```

`across()` pairs naturally with `where()`, `starts_with()`, and the other selection helpers. @sec-iteration-without-loops generalizes the same idea (apply a function to each element of a collection) beyond data frames. But before you build elaborate pipelines out of these verbs, it is worth seeing what they look like without dplyr at all.


### Exercises {.unnumbered}

1. Use `distinct()` to find all unique values of `island`.
2. Use `slice_max()` to find the 5 penguins with the longest flippers.
3. Use `across()` inside `summarise()` to compute the standard deviation (`sd()`) of every numeric column, grouped by species.



## The base R equivalents {#sec-base-r-equivalents}

Everything dplyr does, base R can do too. The verbs are convenient wrappers, not new capabilities. So what exactly do you gain?

`filter()` is logical subsetting:

```{r}
#| eval: false
# dplyr
filter(penguins, species == "Adelie", body_mass_g > 4000)

# base R
penguins[penguins$species == "Adelie" & penguins$body_mass_g > 4000, ]
```

`select()` is column indexing:

```{r}
#| eval: false
# dplyr
select(penguins, species, body_mass_g)

# base R
penguins[, c("species", "body_mass_g")]
```

`mutate()` is direct assignment:

```{r}
#| eval: false
# dplyr
mutate(penguins, mass_kg = body_mass_g / 1000)

# base R
penguins$mass_kg <- penguins$body_mass_g / 1000
```

The base R version modifies the binding; the dplyr version returns a new data frame and leaves the original untouched. This distinction matters more as pipelines grow longer.

`summarise()` plus `group_by()` maps to `aggregate()`:

```{r}
#| eval: false
# dplyr
summarise(group_by(penguins, species), mean_mass = mean(body_mass_g, na.rm = TRUE))

# base R
aggregate(body_mass_g ~ species, data = penguins, FUN = mean, na.rm = TRUE)
```

`arrange()` is `order()`:

```{r}
#| eval: false
# dplyr
arrange(penguins, desc(body_mass_g))

# base R
penguins[order(-penguins$body_mass_g), ]
```

Base R also has `subset()`, which combines filtering and column selection:

```{r}
#| eval: false
subset(penguins, species == "Adelie", select = c(species, body_mass_g))
```

`subset()` is clean and readable; for simple operations, it works well. Where dplyr pulls ahead is composability: every verb takes a data frame and returns one, so you can chain them. Base R's equivalents each return something slightly different: `aggregate()` renames your columns, `order()` gives you indices instead of a data frame. So you spend time wiring them together instead of thinking about your data.

dplyr and SQL\index{SQL} implement the same operations, which is why a dplyr call has a line-for-line SQL translation:

| dplyr | SQL |
|---|---|
| `filter(penguins, species == "Adelie")` | `SELECT * FROM penguins WHERE species = 'Adelie'` |
| `group_by(species) |> summarise(n = n())` | `SELECT species, COUNT(*) AS n ... GROUP BY species` |

Call `show_query()` on a database-backed tibble to see the translation directly.

::: {.callout-note collapse="true"}
## Verb composition is a monoid

A dplyr pipeline is function composition: each verb transforms a data frame and passes it to the next, the same composition from lambda calculus, (g ∘ f)(x) = g(f(x)), applied to tables instead of numbers. This is the same monoid pattern from @sec-no-scalars: an associative operation (verb composition) with an identity element (a verb that returns the data frame unchanged). Compose any sequence of dplyr verbs and the result is still a data frame.
:::

::: callout-tip
## Opinion
Learn the base R equivalents. You will encounter them in other people's code, in Stack Overflow answers, in older packages. But write dplyr in your own work, because the verbs compose, the names communicate intent, and the argument order is consistent. When you read `filter(penguins, species == "Adelie")`, the operation is the function name. When you read `penguins[penguins$species == "Adelie", ]`, you have to parse the indexing to understand what's happening. That parsing cost, multiplied across every line of a script, is exactly the kind of friction Codd's algebra was designed to remove.
:::

Now you have every verb you need to answer the question from the start of the chapter:

```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarise(mean_mass = mean(body_mass_g))
```

Gentoo penguins average roughly 5,076 grams; Adelie average 3,701. The answer was always in the data. Getting to it required four verbs and five lines.


### Exercises {.unnumbered}

1. Rewrite `select(penguins, species, island, body_mass_g)` using base R bracket notation.
2. Rewrite `arrange(penguins, bill_length_mm)` using `order()`.
3. Use `aggregate()` to compute the mean `flipper_length_mm` per island. Compare with the `group_by()` plus `summarise()` version.
