# Pipes and composition {#sec-pipes-and-composition}

```{r}
#| include: false
source(here::here("_common.R"))
```

Consider a line of code that selects columns, filters rows, and sorts the result. Written with nested function calls, you read it inside-out: the first operation you want is buried deepest, and by the third level of parentheses you are counting closing parens by hand. The order you write the code runs backwards from the order you think about the problem.


## The problem with nesting {#sec-nesting-problem}

Real analysis chains multiple operations. Take the `penguins` data, keep only Adelie penguins, select two columns, sort by body mass. Without pipes, you have three choices.

You can nest the calls:

```{r}
#| eval: false
arrange(filter(select(penguins, species, body_mass_g), species == "Adelie"), body_mass_g)
```

This works for two functions, but at three it strains, and at five you are matching parentheses like a Lisp programmer on a bad day. The first operation you want to happen (`select`) is buried deepest inside the expression.

You can name every step:

```{r}
#| eval: false
a <- select(penguins, species, body_mass_g)
b <- filter(a, species == "Adelie")
c <- arrange(b, body_mass_g)
```

Readable, but the environment fills with throwaway names that exist only to shuttle a value from one line to the next: `a`, `b`, `c`, `temp`, `df2`.

Or you can reuse one name:

```{r}
#| eval: false
df <- select(penguins, species, body_mass_g)
df <- filter(df, species == "Adelie")
df <- arrange(df, body_mass_g)
```

Compact, but fragile. You cannot re-run a single line without re-running everything above it, because each line depends on whatever `df` was a moment ago.

All three approaches work, yet none compose well. What would it look like if each step simply fed into the next?


## The pipe: `|>` {#sec-pipe}

```{r}
library(palmerpenguins)
library(dplyr)

penguins |>
  select(species, body_mass_g) |>
  filter(species == "Adelie") |>
  arrange(body_mass_g)
```

Read it top to bottom: take `penguins`, select two columns, keep Adelie rows, sort by mass. Each line transforms the result of the previous one, and the reading order matches the execution order.

`x |> f()`\index{pipe}\index{native pipe} is syntactic sugar for `f(x)`; R rewrites it before the code runs. Chain two and `x |> f() |> g()` becomes `g(f(x))`, the nested form from the previous section written in the order the calls happen.

::: {.callout-note collapse="true"}
## The pipe is a monoid

Chaining functions this way is composition, (g ∘ f)(x) = g(f(x)), and composition is another monoid (@sec-data-transformation): `∘` is the operation, the identity function `\(x) x` is the identity element, and chaining any number of verbs always gives you back a data frame. The same structure shows up again with `Reduce()` in @sec-reduce-and-accumulate.
:::

The native pipe `|>` was introduced in R 4.1 (2021) and is built into the language, requiring no packages. The idea is older than R itself. In 1964 a researcher at Bell Labs wrote a memo about how programs ought to fit together, and one line of it asked for "some ways of coupling programs like garden hose", screwing in another segment whenever the data needed massaging in another way. Unix got that coupling nine years later as the shell's `|` operator, where each program reads from standard input and writes to standard output. The memo was Doug McIlroy's\index{McIlroy, Doug}, and R's `|>` is his garden hose applied to data frames: each dplyr verb takes a data frame and returns one.

::: callout-tip
## Opinion
One verb per line. Put a space before `|>` and a newline after it. A pipeline crammed into a single long line throws away all the readability the pipe was designed to give you.
:::


### Exercises {.unnumbered}

1. Rewrite the following nested call as a pipeline: `head(sort(sqrt(1:20)), 5)`.
2. Take `penguins`, filter for penguins heavier than 4000g, and select only `species` and `island`. Write it as a pipeline.
3. What does `10 |> sqrt()` return? What about `10 |> log(base = 2)`? Explain why `log` receives two arguments.


## Building a pipeline {#sec-building-pipeline}

The best way to build a pipeline is the way you would give directions: one step at a time, checking the view at each turn.

```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarise(mean_mass = mean(body_mass_g)) |>
  arrange(desc(mean_mass))
```

"Take penguins, remove NAs, group by species, compute mean mass, sort descending." The code reads like the sentence.

Build incrementally: run the first line, check the output, add the next step, check again.

```{r}
penguins |>
  filter(!is.na(flipper_length_mm), !is.na(body_mass_g)) |>
  group_by(species, island) |>
  summarise(
    n = n(),
    mean_flipper = mean(flipper_length_mm),
    mean_mass = mean(body_mass_g),
    .groups = "drop"
  )
```

You compose verbs into a pipeline the same way you compose sentences into a paragraph, one idea at a time, in order. But what happens when you want to reuse a pipeline across different data sets, or embed one inside a larger computation?



### Exercises {.unnumbered}

1. Build a pipeline that counts how many penguins of each species live on each island. Start with `penguins`, group by `species` and `island`, then summarise with `n = n()`.
2. Extend your pipeline from Exercise 1: arrange the result by `n` in descending order.
3. Write a pipeline that computes the median `bill_length_mm` per species (removing NAs), then filters to keep only species where the median exceeds 40mm.


## Function composition {#sec-function-composition}

```{r}
clean_and_summarise <- function(df) {
  df |>
    filter(!is.na(body_mass_g)) |>
    group_by(species) |>
    summarise(mean_mass = mean(body_mass_g), .groups = "drop")
}

penguins |> clean_and_summarise()
```

Wrapping a pipeline\index{pipeline} in a function gives the chain a name. `clean_and_summarise` is one function built from three steps, and you can pass it around, store it in a list, or use it inside another pipeline. Applying one function and then another to its result is *function composition*\index{function composition}, the (g ∘ f)(x) = g(f(x)) of @sec-church, and the pipe is R's notation for it.

This is why dplyr verbs all follow the same contract: take a data frame as the first argument, return a data frame. Functions that follow this contract compose via the pipe; functions that don't (first argument is a formula, a filename, a model object) require workarounds. The convention exists so that verbs snap together like sections of garden hose.

The pipeline describes the transformation without naming intermediates:

```r
penguins |> filter(species == "Adelie") |> nrow()
```

::: {.callout-note collapse="true"}
## Point-free style

In Haskell, this style is called "point-free"\index{point-free} or "tacit" programming, tracing back to combinatory logic (Schönfinkel\index{Sch\"{o}nfinkel, Moses}, 1924): functions are composed without mentioning the data they act on. R's pipe enables a mild version of this. It isn't fully point-free (dplyr verbs still reference column names), but the same impulse is at work.
:::


## The placeholder `_` {#sec-placeholder}

The pipe inserts the left side as the first argument. But what happens when the data belongs somewhere else?

```{r}
penguins |>
  filter(!is.na(body_mass_g), !is.na(bill_length_mm)) |>
  lm(body_mass_g ~ bill_length_mm, data = _)
```

The `_` placeholder\index{placeholder} says "put the piped value here." It must appear as a named argument (`data = _`), and it can appear only once. This syntax requires R 4.2+.

For anything more complex, use an anonymous function:

```{r}
c(1, 4, 9, 16) |>
  (\(x) x[x > 3])()
```

The `\(x)` syntax from @sec-anonymous-functions creates a function inline, and the trailing `()` calls it immediately with the piped value. This works for any situation where `_` is too restrictive.


### Exercises {.unnumbered}

1. Pipe `mtcars` into `lm()` to fit `mpg ~ wt`, using the `_` placeholder for the `data` argument.
2. Rewrite the same call using an anonymous function instead of `_`.
3. Why does `penguins |> lm(body_mass_g ~ species, _)` fail? (Hint: the placeholder must be a *named* argument.)


## `|>` vs `%>%` {#sec-pipe-comparison}

The magrittr\index{magrittr (package)} pipe `%>%`\index{\%>\% (magrittr pipe)} came first (2014). The native pipe `|>` arrived in R 4.1 (2021). Both do the same basic thing: insert the left side as the first argument of the right side. The differences are practical.

The placeholder is the difference that matters most: `|>` uses `_` as a named argument, once, while `%>%` uses `.` anywhere and any number of times. The rest are smaller. `|>` requires parentheses on the right side (`x |> sqrt()`, not `x |> sqrt`) and needs no package, while `%>%` needs magrittr (loaded automatically by the tidyverse). The native pipe is also faster: the parser rewrites `x |> f()` to `f(x)` before evaluation, whereas `%>%` is a function call carrying its own overhead.

The `.` pronoun can appear anywhere in the call and more than once, so you can branch on the piped value or pass it to a non-first argument. The following works with `%>%` and has no direct equivalent with `|>`:

```r
x %>% { if (nrow(.) > 0) filter(., cond) else . }
```

`|>` applies the next function. `%>%` applies the next function and also gives you a name for the value you are working with, which makes it more flexible and more complex, and the base pipe leaves that out.

::: {.callout-note collapse="true"}
## Kleisli composition

The base pipe `|>` is plain composition: (g ∘ f)(x), with no implicit variable and no branching. The magrittr pipe `%>%` is reminiscent of what Haskell calls *Kleisli composition*, where you chain functions of the form a -> m b and m is some context (a list, a maybe, an I/O action). The analogy is loose, since `%>%` carries no monadic structure, no `return`, and no bind that satisfies the monad laws, but the shape rhymes: the `.` pronoun threads the result through, and the block above is the kind of branching on the context that Kleisli arrows allow.
:::

Either way, pipes make the order of substitution visible. `h(g(f(x)))` and `x |> f() |> g() |> h()` are the same reductions; the first has to be read from the innermost call outward, the second in the order the functions run. That alignment between notation and execution is why pipes spread through R code within a few years of magrittr's release.

```{r}
# These are equivalent
penguins |> nrow()

library(magrittr)
penguins %>% nrow()
```

::: callout-tip
## Opinion
Use `|>`. It lives in base R, it is faster, and it is the future of the language. The only reason to reach for `%>%` is legacy code or the rare case where you need `.` in multiple positions within a single call.
:::



## When not to pipe {#sec-when-not-to-pipe}

Pipes are for linear sequences: A then B then C. Not every problem is linear.

If two data frames need to interact, pipe one and pass the other as an argument. A join fits naturally in a pipeline; a complex merge of three data frames, each requiring its own preparation steps, does not.

If you need intermediate results for different purposes later, name them. A pipeline produces one result, so if you need both the filtered data and the summary, use two separate pipelines from a shared intermediate.

```{r}
adelie <- penguins |> filter(species == "Adelie")

# Two different analyses from the same intermediate
adelie |> summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))
adelie |> count(island)
```

Past eight or ten steps, break the pipeline and name the intermediate result at a meaningful boundary. A pipeline that scrolls off the screen is hard to read and harder to debug.

`print()`, `write.csv()`, and `plot()` produce side effects; they belong at the end of a pipeline, not wedged between two transformations where they interrupt the data flow.

When a pipeline produces unexpected output, break it apart. Assign intermediates, inspect each one, find the step that goes wrong, fix it, and reassemble the pipeline.

::: callout-tip
## Opinion
A pipe should tell a story. If it reads like a run-on sentence, break it into paragraphs and name the paragraph boundaries with meaningful variable names, not `temp1` and `temp2`.
:::

A pipeline applies a chain of transformations to one data frame. Applying the same chain to every element of a list is the job of `map()`, in @sec-purrr-map.

### Exercises {.unnumbered}

1. The following pipeline tries to do too much. Break it into two or three named steps with meaningful names:

```{r}
#| eval: false
penguins |>
  filter(!is.na(body_mass_g)) |>
  mutate(mass_kg = body_mass_g / 1000) |>
  group_by(species) |>
  summarise(mean_kg = mean(mass_kg)) |>
  arrange(desc(mean_kg)) |>
  mutate(label = paste(species, round(mean_kg, 1), "kg"))
```

2. Why is putting `write.csv()` in the middle of a pipeline (between `filter` and `summarise`) a bad idea? What could go wrong?
3. You need both a summary table (mean mass per species) and a filtered dataset (Adelie penguins only) from `penguins`. Write code that avoids repeating the `filter(!is.na(body_mass_g))` step.
