# Tidy data {#sec-tidy-data}

```{r}
#| include: false
source(here::here("_common.R"))
```

You have a data frame with population figures for three countries, and the years are column names: `pop_2020`, `pop_2021`, `pop_2022`. You want to plot population over time, so you try mapping `year` to the x-axis. There is no `year` column. The information is there, hiding in plain sight, encoded in the syntax of your column headers rather than the semantics of your data. Every dplyr verb you learned in @sec-data-transformation, every ggplot2 aesthetic you'll ever write, every model formula you'll ever fit: they all assume that each variable lives in its own column. When it doesn't, nothing works. Getting data into the right shape occupies more of a working analyst's time than any model or visualization, and yet most introductions to R treat it as an afterthought, something you do before the real work begins. This chapter treats it as the real work, because it is.

You already have the pieces: data frames from @sec-data-frames, verbs from @sec-data-transformation, pipes from @sec-pipes-and-composition. What you're missing is the target shape those tools expect.


## What tidy means {#sec-what-tidy-means}

Hadley Wickham's 2014 paper "Tidy data" defines three rules:

1. Each variable is a column.
2. Each observation is a row.
3. Each type of observational unit is a table.

They sound obvious. They aren't. Most real-world data breaks at least one of them, and recognizing which rule is broken (let alone fixing it) takes practice. The rules themselves are a simplified version of something database designers have known since the 1970s, when Edgar Codd defined normal forms (1NF through 5NF) to eliminate redundancy; tidy data corresponds roughly to third normal form, with no transitive dependencies and no repeating groups. The structural problem R programmers wrestle with today is the same one Codd solved in relational algebra before R existed.

Consider a table of population data with years spread across the column names:

```{r}
pop_wide <- data.frame(
  country    = c("Norway", "Sweden", "Denmark"),
  pop_2020   = c(5380, 10350, 5831),
  pop_2021   = c(5391, 10380, 5840),
  pop_2022   = c(5408, 10420, 5857)
)
pop_wide
```

What is the variable here? Year. But "year" doesn't appear as a column; it's buried inside the names `pop_2020`, `pop_2021`, `pop_2022`, three columns where there should be one. Not tidy.

The tidy version has three columns: `country`, `year`, and `population`. One variable per column, one observation per row, nine values arranged so that every tool you've learned can reach them.

```{r}
pop_tidy <- data.frame(
  country    = rep(c("Norway", "Sweden", "Denmark"), each = 3),
  year       = rep(2020:2022, times = 3),
  population = c(5380, 5391, 5408, 10350, 10380, 10420, 5831, 5840, 5857)
)
pop_tidy
```

Now you can filter by year, group by country, plot year on the x-axis. Every dplyr verb and every ggplot2 aesthetic mapping works because each variable occupies its own column, and the tools compose because the data has a consistent shape. But who wants to rearrange nine values by hand?


### Exercises {.unnumbered}

1. Look at this table. Which tidy-data rule does it break?

```{r}
#| eval: false
data.frame(
  student = c("Alice", "Bob"),
  math_score = c(90, 85),
  english_score = c(78, 92)
)
```

2. Sketch (on paper or in a comment) what the tidy version of that table would look like. How many rows? How many columns?

3. A data frame has columns `name`, `date`, and `temperature`. One row per city per day. Is this tidy? Why or why not?


## `pivot_longer()`: columns to rows {#sec-pivot-longer}

`pivot_longer()` from the tidyr package does the reshaping in one call:

```{r}
library(tidyr)
library(dplyr)

pop_wide |>
  pivot_longer(
    cols      = pop_2020:pop_2022,
    names_to  = "year",
    values_to = "population"
  )
```

Three countries, three years, nine population values: the information content is identical in both forms. `pivot_longer()` and `pivot_wider()` are inverses; applying one and then the other returns you to where you started (up to column ordering). A pair of reversible transformations like this is called an *isomorphism*, two structures that contain the same information in different arrangements. Wide and long formats are isomorphic; neither is "more correct." In category theory, this kind of structure-preserving, reversible map between containers is called a *natural transformation*: it rearranges data inside the data-frame container without changing the information content, the same way `map()` transforms values inside a list without changing the list's length.

The hard part is not the reshaping itself but deciding *what* to reshape. Given an arbitrary table with merged cells, hierarchical headers, and embedded metadata, determining the correct tidy structure requires human judgment about what constitutes a variable, an observation, and a value. Wickham's 2014 paper defines these three terms precisely to make that judgment tractable: once you answer "is year a variable or a value?", the reshaping follows mechanically.

Three arguments control the transformation:

- `cols`: which columns to pivot. Accepts tidy selection helpers like `starts_with()`, `where(is.numeric)`, or a range with `:`.
- `names_to`: what to call the new column made from column names.
- `values_to`: what to call the new column made from cell values.

The column names still carry the `pop_` prefix, and `year` is character, not integer. Two more arguments fix that:

```{r}
pop_wide |>
  pivot_longer(
    cols            = pop_2020:pop_2022,
    names_to        = "year",
    names_prefix    = "pop_",
    names_transform = list(year = as.integer),
    values_to       = "population"
  )
```

`names_prefix` strips the leading text; `names_transform` applies a function to the extracted names, here converting character years to integers.

For more complex column names, `names_sep` and `names_pattern` let you split structured names into multiple new columns. If you had columns like `bill_length_mm` and `bill_depth_mm`, you could split on `_` to extract both the measurement and the unit. The tidyr pivoting vignette covers these cases in detail. But what about the opposite direction, when your data is *too* long?


### Exercises {.unnumbered}

1. Create this data frame and pivot it to tidy form with columns `city`, `quarter`, and `sales`:

```{r}
#| eval: false
quarterly <- data.frame(
  city = c("Oslo", "Bergen"),
  Q1   = c(120, 95),
  Q2   = c(135, 110),
  Q3   = c(150, 105),
  Q4   = c(160, 130)
)
```

2. Modify your pivot to strip the `Q` prefix from the quarter column and convert it to an integer.

3. What happens if you set `values_to` to the name of an existing column? Try it.


## `pivot_wider()`: rows to columns {#sec-pivot-wider}

Sometimes the problem goes the other way: distinct variables are stacked into a single column, and you need to spread them out.

```{r}
measurements <- data.frame(
  site        = rep(c("A", "B"), each = 2),
  measurement = rep(c("temperature", "humidity"), times = 2),
  value       = c(22.1, 65, 19.8, 72)
)
measurements
```

Temperature and humidity are different variables, and cramming them into a single `value` column conflates two things that should be separate. `pivot_wider()` pulls them apart:

```{r}
measurements |>
  pivot_wider(
    names_from  = measurement,
    values_from = value
  )
```

Two arguments:

- `names_from`: which column's values become the new column names.
- `values_from`: which column's values fill the new cells.

When some combinations don't exist in the data, `pivot_wider()` fills those cells with `NA`. The `values_fill` argument lets you choose a different default:

```{r}
incomplete <- data.frame(
  site        = c("A", "A", "B"),
  measurement = c("temperature", "humidity", "temperature"),
  value       = c(22.1, 65, 19.8)
)

incomplete |>
  pivot_wider(
    names_from  = measurement,
    values_from = value,
    values_fill = 0
  )
```

::: callout-tip
## Opinion
If you reach for `pivot_wider()` during analysis, pause and ask whether your downstream tool really needs wide data. Usually it doesn't. `pivot_wider()` is most useful for building display tables and for feeding data into functions that expect matrix-like input. The fact that wide format *looks* tidier to human eyes doesn't mean it *is* tidier for your code.
:::


### Exercises {.unnumbered}

1. Start with `pop_tidy` (the tidy population table from @sec-what-tidy-means). Use `pivot_wider()` to recreate the wide format, with one column per year.

2. What happens if the combination of id columns isn't unique before widening? Create a data frame where two rows have the same `site` and `measurement` and try `pivot_wider()`. Read the warning.


## When data isn't tidy (and that's fine) {#sec-when-not-tidy}

Tidy data is for analysis. Not every use case is analysis.

Wide tables are easier for humans to scan: a table with countries as rows and years as columns fits on a screen, while the same data in long form might scroll for hundreds of rows. Wide format is also what some matrix operations, correlation computations, and model interfaces expect when they want one column per predictor.

Data entry follows the same logic. If you're recording daily temperatures for three cities, a spreadsheet with cities as rows and dates as columns feels natural; nobody fills in a three-column long-form table by hand (and anyone who tells you they do is lying, or masochistic, or both).

The workflow is: receive data however it arrives, reshape to tidy for analysis, reshape to wide for presentation if needed. Tidy data is a means to easier analysis, not a moral stance.

But what happens when the information you need spans multiple tables?


## Joins: combining tables {#sec-joins}

You have customers and orders, species and observations, sites and measurements. The data lives in separate tables, linked by a shared key. Joins are how you bring them together.

```{r}
bands <- data.frame(
  name  = c("John", "Paul", "George", "Ringo"),
  plays = c("guitar", "bass", "guitar", "drums")
)

albums <- data.frame(
  name  = c("John", "Paul", "George", "Mick"),
  album = c("Imagine", "Ram", "All Things Must Pass", "Some Girls")
)
```

### Mutating joins {.unnumbered}

Mutating joins add columns from the second table to the first.

`left_join()` keeps all rows from the left table and attaches matching columns from the right, filling `NA` where no match exists:

```{r}
bands |> left_join(albums, by = "name")
```

Ringo has no album in the `albums` table, so his `album` comes back `NA`; Mick has no row in `bands`, so he's dropped entirely. This is the workhorse join: keep everything on the left, enrich with what you can from the right.

`inner_join()` is stricter, keeping only rows that match in both tables:

```{r}
bands |> inner_join(albums, by = "name")
```

Ringo and Mick are both gone. Only the three names appearing in both tables survive.

When the key columns have different names, use `join_by()`:

```{r}
instruments <- data.frame(
  musician   = c("John", "Paul", "George", "Ringo"),
  instrument = c("guitar", "bass", "guitar", "drums")
)

albums |> left_join(instruments, by = join_by(name == musician))
```

### Filtering joins {.unnumbered}

Filtering joins don't add columns. They filter rows.

`semi_join()` keeps rows in the left table that have a match in the right:

```{r}
bands |> semi_join(albums, by = "name")
```

Same three rows as `inner_join()`, but without the `album` column. You're using `albums` as a filter, not as a source of new data.

`anti_join()` does the opposite, keeping rows in the left table that have *no* match in the right:

```{r}
bands |> anti_join(albums, by = "name")
```

Only Ringo: the one band member with no album in the `albums` table. `anti_join()` is your best tool for finding what's missing. Which customers placed no orders? Which species have no observations? Which students didn't submit?

::: callout-tip
## Opinion
Start with `left_join()`. Use `anti_join()` to debug. The others are situational. `right_join()` exists but is rarely needed: swap the table order and use `left_join()` instead. `full_join()` keeps everything from both sides, which sounds safe but usually signals that you haven't decided what your unit of analysis actually is.
:::


### Exercises {.unnumbered}

1. Create two small data frames: `students` with columns `id` and `name`, and `grades` with columns `id` and `score`. Give them overlapping but not identical `id` values. Use `left_join()` to combine them.

2. Use `anti_join()` on the same tables to find students with no grades and grades with no student. (Hint: swap which table is on the left.)

3. What is the difference between `semi_join(bands, albums, by = "name")` and `inner_join(bands, albums, by = "name") |> select(name, plays)`? Try both.


## Other tidyr tools {#sec-other-tidyr}

Pivoting handles structure; joins handle combination. A handful of other tidyr functions handle the messy details that remain after those two big moves.

### Separating and uniting columns {.unnumbered}

`separate_wider_delim()` splits one column into several by a delimiter:

```{r}
dates <- data.frame(
  id   = 1:3,
  date = c("2024-03-15", "2024-07-22", "2024-11-01")
)

dates |>
  separate_wider_delim(
    cols  = date,
    delim = "-",
    names = c("year", "month", "day")
  )
```

This replaces the older `separate()`, which relied on position-based splitting and was less predictable. For complex patterns, `separate_wider_regex()` lets you use regular expressions.

`unite()` does the inverse, pasting columns together:

```{r}
data.frame(year = 2024, month = "03", day = 15) |>
  unite(col = "date", year, month, day, sep = "-")
```

### Completing and filling {.unnumbered}

`complete()` makes implicit missing values explicit. If your data has observations for some year/site combinations but not all, `complete()` adds rows for the gaps and fills them with `NA`:

```{r}
obs <- data.frame(
  year = c(2020, 2020, 2021),
  site = c("A", "B", "A"),
  count = c(5, 3, 7)
)

obs |> complete(year, site)
```

`fill()` propagates the last non-missing value downward (or upward), which is useful for data where group labels are only written once and the rows beneath are left blank:

```{r}
report <- data.frame(
  group = c("Control", NA, NA, "Treatment", NA),
  value = c(10, 12, 11, 20, 22)
)

report |> fill(group)
```

### Nesting {.unnumbered}

`nest()` packs subsets of a data frame into list-columns, and `unnest()` unpacks them. This pattern shines when fitting models to groups, but it requires familiarity with list-columns and iteration, both covered in @sec-purrr-map.


## Summary {#sec-tidy-data-summary}

Tidy data is a structural contract: one variable per column, one observation per row, one observational unit per table. `pivot_longer()` turns columns into rows (the common direction), `pivot_wider()` turns rows into columns (the occasional direction), and joins combine tables by matching keys. `left_join()` enriches; `anti_join()` finds gaps; the rest are variations on the same theme.

These tools compose precisely because they all expect the same shape, and the connection to their mathematical foundations runs deeper than most R programmers realize. Codd's relational algebra and the simply typed lambda calculus are both built on function application and composition over structured types, which is why the same ideas keep surfacing in R, SQL, and formal logic. In category theory, a join is a *pullback*: given two tables that each map to a shared key space, the join constructs the most general table consistent with both. Think of it as a Venn diagram for rows, where the overlap is determined by the key column; `inner_join` keeps only the overlap, while `left_join` keeps everything on the left and fills the gaps with `NA`.

But shape is only half the story. Once your data is tidy, you still need to *do* something with it, and doing the same thing to many groups, many columns, or many models at once is where iteration enters. That's next.
