# Iteration without loops {#sec-iteration-without-loops}

```{r}
#| include: false
source(here::here("_common.R"))
library(palmerpenguins)
library(purrr)
library(dplyr)
```

You have a list of three data frames and you want the mean of one column in each. You can write six lines of `for` loop, or you can hand the computation to a function that does the looping for you. @sec-passing-functions showed functions being passed as arguments; this chapter is what that buys you.

This chapter uses three packages. Load them now if you are following along:

```{r}
#| message: false
library(palmerpenguins)
library(purrr)
library(dplyr)
```

We will work with the penguins data split by species throughout:

```{r}
penguins_split <- split(penguins, penguins$species)
names(penguins_split)
```

Three data frames, one per species. Almost everything in this chapter amounts to doing something to each of them, and the question is always the same: how much of the work should you spell out yourself, and how much should the language handle?


## The problem with `for` loops {#sec-problem-with-for-loops}

Suppose you want the mean body mass for each species. Here is the loop version:

```{r}
results <- vector("list", length(penguins_split))
for (i in seq_along(penguins_split)) {
  results[[i]] <- mean(penguins_split[[i]]$body_mass_g, na.rm = TRUE)
}
names(results) <- names(penguins_split)
results
```

Six lines to say "compute the mean mass for each group."\index{for loop} Look at how much of that is bookkeeping: pre-allocate a container, set up an index, store results at the right position, copy names over at the end. The actual computation, `mean(df$body_mass_g, na.rm = TRUE)`, sits buried in the middle, surrounded by scaffolding that has nothing to do with the question you asked.

A functional says the same thing in one line:

```{r}
map_dbl(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE))
```

Same answer. No index variable, no pre-allocation, no name copying. You said what to compute and `map_dbl()` handled the iteration\index{iteration}. A function that takes another function and applies it across a collection like this is called a *functional*\index{functional}.

A well-written loop in modern R runs as fast as the functional. What the loop mixed together was the *mechanism* of iteration (allocate, index, store) and the *content* of the computation (compute a mean). The functional keeps the mechanism inside itself and leaves the content in the function you pass, and once the two are apart, each is easier to read, test, and reuse on its own.

::: callout-tip
## Opinion
For loops are verbose R. Once the same scaffolding appears three or four times in a project, a functional replaces all of it with one line.
:::

### Exercises {.unnumbered}

1. Write a `for` loop that computes the number of rows in each element of `penguins_split`. Then rewrite it using `map_int()` and `nrow`.
2. What happens if you forget to pre-allocate `results` and instead grow it with `results <- c(results, value)` inside a loop? Why is this slower?


## `lapply()` and `sapply()`: base R functionals {#sec-lapply-sapply}

`lapply()`\index{lapply()} is base R's version of the same idea:

```{r}
lapply(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE))
```

`lapply(x, f)` applies `f` to each element of `x` and returns a list. Always a list. That predictability is its strength.

::: {.callout-note collapse="true"}
## Where `map` comes from

Handing a function to something that applies it to every element of a list is older than R and older than S. John McCarthy's 1960 paper on Lisp defined `maplist`, which applies a function to each successive tail of a list, and the LISP 1.5 manual added `mapcar`, which applies it to each element. Later languages (Scheme, ML, Haskell) settled on the name `map`. R inherited the idea twice: once as `lapply()` in base R, and again as `purrr::map()` in the tidyverse.
:::

`sapply()`\index{sapply()} does the same thing, but tries to "simplify" the result:

```{r}
sapply(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE))
```

When every call to `f` returns a single value, `sapply()` hands you a named vector. But if `f` returns different lengths for different elements, `sapply()` falls back to a list. You cannot predict the return type without knowing the output of `f` for every element of `x`. Fine at the console; unreliable inside a function that other code depends on.

`vapply()`\index{vapply()} solves this by making you declare the expected type:

```{r}
vapply(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE), numeric(1))
```

The third argument, `numeric(1)`, says "I expect each call to return a single numeric value." If any call returns something else, you get an error instead of a silent type change.

Two other base R functionals are worth knowing.\index{apply family} `tapply(x, group, f)` applies `f` to subsets of `x` defined by `group`:

```{r}
tapply(penguins$body_mass_g, penguins$species, mean, na.rm = TRUE)
```

And `mapply(f, x, y)` iterates over multiple inputs in parallel:

```{r}
mapply(paste, c("one", "two", "three"), c("fish", "cat", "bird"))
```

::: callout-tip
## Opinion
Use `lapply()` when you want a list. Use `vapply()` when you want a vector and want safety. Never use `sapply()` in non-interactive code: its return type depends on the data, and that is a bug waiting in your pipeline.
:::

### Exercises {.unnumbered}

1. Use `lapply()` and `nrow` to get the number of rows in each element of `penguins_split`.
2. Use `vapply()` to compute the median flipper length for each species. Declare the expected output type.
3. What does `sapply(list(), mean)` return? What about `vapply(list(), mean, numeric(1))`? Which is safer?


## `purrr::map()`: the tidyverse functional {#sec-purrr-map}

The purrr\index{purrr (package)} package provides `map()`\index{map()}, which does the same thing as `lapply()`:

```{r}
map(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE))
```

Apply a function to each element, return a list. So why bother with a new package?

The typed variants. `map_dbl()`\index{map\_dbl()} returns a double vector, `map_chr()`\index{map\_chr()} a character vector, and if the function returns the wrong type you get an error where `sapply()` would have handed you a list:

```{r}
map_dbl(penguins_split, \(df) mean(df$body_mass_g, na.rm = TRUE))
```

Like `lapply()`, `map()` takes the collection first, so it sits naturally after a pipe:

```{r}
penguins_split |>
  map_dbl(\(df) mean(df$body_mass_g, na.rm = TRUE))
```

You may encounter the formula shorthand `~ mean(.x$body_mass_g, na.rm = TRUE)` in older code, purrr's anonymous-function syntax from before R 4.1 introduced `\(x)`. Prefer `\(x)` in new code.

`map()` always returns a list and `map_dbl()` always returns a double vector, whatever the data. In a long pipeline that is what lets you know the shape of every intermediate result without running it. And the list has three elements, so the function runs three times: the structure of the data determines how many applications occur, and you stop managing the iteration yourself.

::: {.callout-note collapse="true"}
## Functors and the shape of iteration

`map()` embodies a pattern from category theory: the *functor*. `map(xs, f)` applies `f` to each element of `xs`, preserving the container structure (a list of three elements in, a list of three elements out). A functor lifts a function between values (`f :: A -> B`) to a function between containers (`map(f) :: List(A) -> List(B)`). You do not need the theory to use `map()`, but recognizing the pattern explains why `map()` feels the same whether you apply it to a list, a vector, or a set of data frame columns with `across()`: it is the same structural idea in different containers.
:::

### Exercises {.unnumbered}

1. Use `map_dbl()` to compute the standard deviation of `body_mass_g` for each species in `penguins_split`.
2. Use `map_chr()` to extract the first value of `species` from each element of `penguins_split`. (Hint: `\(df) as.character(df$species[1])`.)
3. Rewrite the `for` loop from @sec-problem-with-for-loops using `map_dbl()`. Compare the two versions for readability.


## Typed variants {#sec-typed-variants}

`map_int()` and `map_lgl()` complete the set:

```{r}
map_int(penguins_split, nrow)
```

```{r}
map_lgl(penguins_split, \(df) any(is.na(df$body_mass_g)))
```

Each typed variant checks both the type and the length of every result: `map_dbl()` wants a single double per call, `map_chr()` a single string, `map_lgl()`\index{map\_lgl()} a single logical, `map_int()`\index{map\_int()} a single integer. Anything else is an error:

```{r}
#| error: true
map_dbl(penguins_split, \(df) df$body_mass_g)
```

When the output does not match your expectation, you find out immediately, not three functions downstream when something produces `NULL` instead of a number.

For results that are data frames, the current pattern is `map()` followed by `list_rbind()`; the older `map_dfr()` still works but is superseded:

```{r}
penguins_split |>
  map(\(df) tibble(
    species = df$species[1],
    mean_mass = mean(df$body_mass_g, na.rm = TRUE),
    n = nrow(df)
  )) |>
  list_rbind()
```

`map()` returns a list of data frames, and `list_rbind()` stacks them into one. But what about iterating over more than one thing at a time?

### Exercises {.unnumbered}

1. Use `map_int()` and `nrow` to count the rows in each element of `penguins_split`.
2. Use `map_lgl()` to check which elements of `penguins_split` have more than 100 rows.
3. Use `map()` and `list_rbind()` to build a summary data frame with one row per species, containing the species name, median bill length, and median bill depth.


## Iterating over multiple inputs {#sec-map2-pmap}

Sometimes you need to walk over two inputs in lockstep, pulling one element from each on every step. `map2()` handles this:

```{r}
species_names <- names(penguins_split)
species_counts <- map_int(penguins_split, nrow)

map2_chr(species_names, species_counts, \(name, n) paste(name, ":", n, "penguins"))
```

`map2(.x, .y, .f)`\index{map2()} calls `.f(x[[1]], y[[1]])`, then `.f(x[[2]], y[[2]])`, and so on. The two inputs are consumed in parallel, not in a nested grid.

For three or more inputs, use `pmap()`\index{pmap()}. You supply a list of vectors (or a data frame, since a data frame is a list of columns):

```{r}
params <- list(
  mean = c(0, 5, -3),
  sd = c(1, 2, 0.5),
  n = c(3, 3, 3)
)

set.seed(42)
pmap(params, \(mean, sd, n) rnorm(n, mean, sd))
```

Each row of parameters becomes one call. `pmap()` is the general case; `map2()` is the special case for exactly two inputs.

The base R equivalents are `Map(f, x, y)`\index{Map()} (which returns a list, like `map2()`) and `mapply(f, x, y)` (which simplifies, like `sapply()`). The same warning applies: `mapply` is unpredictable in return type.

`pmap()` earns its place when a table of parameters drives the work: different models on different subsets with different formulas, all in one tibble. Suppose you want to regress body mass on flipper length for each species, but with a different formula for each:

```{r}
configs <- tibble(
  data = penguins_split,
  formula = list(
    body_mass_g ~ flipper_length_mm,
    body_mass_g ~ flipper_length_mm + bill_length_mm,
    body_mass_g ~ flipper_length_mm * bill_length_mm
  )
)

models <- pmap(configs, \(data, formula) lm(formula, data = data))
map_dbl(models, \(m) summary(m)$r.squared)
```

`pmap()` iterates over the rows of the configuration table, fitting one model per species. The entire experiment lives in one tibble, and adding a fourth species or a fourth formula means adding a row, not rewriting the loop.

### Exercises {.unnumbered}

1. Use `map2_chr()` to paste together the vectors `c("one", "two", "three")` and `c("fish", "cat", "bird")` with a space in between.
2. Create a list of three means and three standard deviations. Use `map2_dbl()` to generate one random normal value for each pair. (Hint: `\(m, s) rnorm(1, m, s)`.)
3. A data frame is a list of columns. What does `pmap_dbl(data.frame(x = 1:3, y = 4:6), \(x, y) x + y)` return?


## Side effects: `walk()` {#sec-walk}

Everything so far has been about functions that *compute* something: they take input and produce output. But some functions exist for what they *do*, not what they *return*: printing a message, writing a file, drawing a plot. For these, use `walk()`.

```{r}
walk(names(penguins_split), \(name) cat(name, "\n"))
```

`walk(.x, .f)`\index{walk()} applies `.f` to each element of `.x`, but instead of collecting results into a list, it returns `.x` invisibly. The point is the side effect, not the return value.

The two-input version, `walk2()`, is especially useful for writing files:

```{r}
#| eval: false
file_paths <- paste0("data/", names(penguins_split), ".csv")
walk2(penguins_split, file_paths, \(df, path) write.csv(df, path, row.names = FALSE))
```

Each data frame gets written to its own file. `walk2()` iterates over the data frames and paths in parallel, calling `write.csv()` for each pair.

Use `walk()` whenever `.f` does something rather than computes something. If you use `map()` for side effects, you end up with a list of `NULL` values cluttering your console, which is the language telling you that you reached for the wrong tool.

Because `walk()` returns its input invisibly, it slots cleanly into the middle of a pipeline:

```{r}
penguins_split |>
  walk(\(df) cat("Species:", as.character(df$species[1]), "- n =", nrow(df), "\n")) |>
  map_dbl(\(df) mean(df$body_mass_g, na.rm = TRUE))
```

The `walk()` prints a line for each species, then passes `penguins_split` through unchanged to `map_dbl()`. Side effects and computation in one pipeline. One kind of iteration, though, is still out of reach for every functional so far.

### Exercises {.unnumbered}

1. Use `walk()` to print the column names of each element of `penguins_split`. (They should all be the same.)
2. Use `walk2()` and `cat()` to print lines like `"Adelie has 152 rows"` for each species.


## When loops are fine {#sec-when-loops-are-fine}

Not everything should be a functional. Here is a random walk, where each step depends on the one before it:

```{r}
x <- numeric(20)
x[1] <- 0
for (i in 2:20) {
  x[i] <- x[i - 1] + rnorm(1)
}
x
```

`map()` cannot write this, because it hands each element to the function on its own and assumes the elements are independent. Whenever iteration `i` needs the result of iteration `i - 1` (simulation chains, iterative algorithms, anything that carries state), the loop is the honest tool. purrr's `reduce()` and `accumulate()` handle some of these sequential cases, and @sec-reduce-and-accumulate covers them.

A loop is also clearer when the body is complex. `next` to skip elements, `break` to stop early, several branches that depend on runtime state: contorting all of that into a function passed to `map()` costs more than it saves.

And a loop you understand is better than a functional you do not. Use `map()` when the pattern clicks; until then, write the loop, get the right answer, and refactor later when the shape of the problem becomes obvious.

::: callout-tip
## Opinion
Replace a loop with a functional when the pattern is "do this to each thing independently." Keep a loop when iterations talk to each other.
:::


## `across()` revisited {#sec-across-revisited}

In @sec-more-verbs you used `across()`\index{across()} to apply a function to several columns inside a dplyr verb:

```{r}
penguins |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

`across(.cols, .fns)` takes a column selection and a function (or a list of functions) and applies the function to each selected column. That is `map()` over the columns of a data frame, specialized for the inside of a dplyr verb.

Multiple functions at once:

```{r}
penguins |>
  summarise(across(
    where(is.numeric),
    list(mean = \(x) mean(x, na.rm = TRUE), sd = \(x) sd(x, na.rm = TRUE))
  ))
```

Combined with `group_by()`:

```{r}
penguins |>
  group_by(species) |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

`across()` replaces the older `summarise_at()`, `summarise_if()`, and `summarise_all()`, which are now superseded. If you encounter them in older code, they do the same thing with less flexible syntax.

Under the hood, `across()` iterates over the selected columns, applies the function to each one, and assembles the results. That is the same thing `map()` does over list entries and `sapply()` does over vector values. The `for` loop at the start of this chapter, `lapply()`, `map_dbl()`, `across()`: each separates *what to do* from *how to iterate*, and each walks a different container with the same logic.

### Exercises {.unnumbered}

1. Use `across()` inside `summarise()` to compute the `median()` of every numeric column in `penguins`, removing `NA` values.
2. Use `across()` with a list of two functions to compute both the `min()` and `max()` of every numeric column, grouped by species.
3. What does `across(where(is.character), n_distinct)` compute when used inside `summarise()`? Try it on `penguins`.
