# Function factories {#sec-function-factories}

```{r}
#| include: false
source(here::here("_common.R"))
```

Suppose you need twelve formatting functions: one for dollars, one for euros, one for percentages, one for each of nine other currencies your client cares about. You could write twelve functions by hand, each identical except for a prefix string. Or you could write one function that manufactures the other twelve. You have already built one of these: `make_adder` returns a function that remembers `n`, and @sec-closures-and-scope explained why the memory holds, because the returned function is a closure over the environment it was made in. This chapter builds whole families of functions from a single template.


## The pattern {#sec-factory-pattern}

Start with one that raises numbers to a power:

```{r}
power <- function(exponent) {
  function(x) x ^ exponent
}

square <- power(2)
cube <- power(3)
```

```{r}
square(5)
cube(5)
```

When you call `power(2)`, R creates an execution environment where `exponent` is 2, then returns an anonymous function that closes over that environment. `square` remembers `exponent = 2` forever; `cube` remembers `exponent = 3`. Same logic, different parameters. `power` produces a *family* of related functions, and each member carries its own private copy of the parameter that distinguishes it from its siblings.

A function that returns a function this way is a *function factory*\index{function factory}, and the functions it produces are closures.

::: {.callout-note collapse="true"}
## Partial application and currying

`power(2)` fixes one argument of a two-argument operation and returns a function of the remaining argument: *partial application*\index{partial application}. In lambda calculus, `power` is `λe. λx. x^e`, and applying it to 2 gives `λx. x^2`, the currying from @sec-arguments-defaults. Haskell curries every function by default, so `add 5 3` is really `(add 5) 3`, with `add 5` returning a function. A function factory in R is currying done by hand.
:::


### Exercises {.unnumbered}

1. Write a function factory `make_multiplier` that takes a `factor` and returns a function that multiplies its argument by `factor`. Create `double` and `triple` and test them on the number 7.
2. What does `power(1)` return? Is it the identity function? Test it.
3. Write a factory `make_greeter` that takes a greeting string and returns a function that takes a name and produces the greeting. `make_greeter("Hello")("Alice")` should return `"Hello, Alice"`.


## The lazy evaluation trap {#sec-lazy-eval-trap}

Here is a bug that costs an afternoon the first time you meet it:

```{r}
exp <- 2
sq <- power(exp)
exp <- 3
sq(5)
```

That returns 125, not 25. R evaluates an argument the first time it is used, which can be long after the call, and until then it stores the unevaluated expression in an object called a *promise*\index{promise} (@sec-promises has the full mechanism). Inside `power(exp)`, `exponent` is a promise holding the expression `exp`. Nothing in the factory body uses `exponent`, so it stays a promise, and by the time `sq(5)` finally needs it, `exp` in the calling environment has become 3. The factory captured a promise, not a value. This is the *lazy evaluation trap*\index{lazy evaluation trap}.

The loop version is worse:

```{r}
fns <- list()
for (i in 1:3) {
  fns[[i]] <- function(x) x ^ i
}

fns[[1]](2)
fns[[2]](2)
fns[[3]](2)
```

All three return 8. By the time any of them is called, `i` is 3.

The fix is `force()`\index{force()}:

```{r}
power <- function(exponent) {
  force(exponent)
  function(x) x ^ exponent
}
```

`force(exponent)` evaluates the argument immediately, capturing the value rather than the promise, and now the factory works correctly:

```{r}
exp <- 2
sq <- power(exp)
exp <- 3
sq(5)
```

25, as expected. The loop version has no argument to force: all three functions share the single `i` in the global environment. `lapply()` sidesteps the problem:

```{r}
fns <- lapply(1:3, \(i) function(x) x ^ i)

fns[[1]](2)
fns[[2]](2)
fns[[3]](2)
```

Why does this work when the `for` loop didn't? Because the anonymous function `\(i)` creates a new scope for each iteration. When `lapply` calls `\(i)` with the value `1`, that call gets its own execution environment where `i` is `1`, and the inner `function(x) x ^ i` closes over that environment. The next call gets a fresh environment where `i` is `2`, and so on. Each manufactured function captures a different `i` in a different environment, whereas in the `for` loop there is only one `i` in one environment and every function points to it. `lapply()` also forces the argument before `\(i)` returns, so the promise problem from `power(exp)` cannot recur inside it.

::: callout-tip
## Opinion
Every factory should `force()` every argument that the manufactured function uses. Make this a habit, not a debugging exercise. `force()` costs nothing, and the bug it prevents is invisible from the outside.
:::


### Exercises {.unnumbered}

1. Predict the output of the following code, then run it to check:

    ```r
    val <- 10
    make_adder <- function(n) function(x) x + n
    add_val <- make_adder(val)
    val <- 20
    add_val(1)
    ```

2. Fix the `make_adder` factory above with `force()`. Verify that changing `val` after creating `add_val` no longer affects the result.
3. Why does `lapply(1:3, \(i) function(x) x ^ i)` work without an explicit `force()` call?


## Practical factories {#sec-practical-factories}

The pattern shows up wherever the same logic repeats with different parameters. A formatter, for one, can have its prefix and suffix baked in:

```{r}
make_formatter <- function(prefix, suffix = "") {
  force(prefix)
  force(suffix)
  function(x) paste0(prefix, x, suffix)
}

usd <- make_formatter("$")
pct <- make_formatter("", "%")
```

```{r}
usd(42)
pct(0.15)
```

A filter can carry its cutoff:

```{r}
above <- function(threshold) {
  force(threshold)
  function(x) x[x > threshold]
}

above_zero <- above(0)
above_zero(c(-2, 0, 3, -1, 5))
```

And a statistical transformation that depends on a single parameter is a natural factory. The Box-Cox transformation is `log(x)` when its parameter lambda is 0 and `(x^lambda - 1) / lambda` otherwise:

```{r}
box_cox <- function(lambda) {
  force(lambda)
  if (lambda == 0) {
    \(x) log(x)
  } else {
    \(x) (x ^ lambda - 1) / lambda
  }
}

bc1 <- box_cox(1)
bc0 <- box_cox(0)

bc1(c(1, 2, 4))
bc0(c(1, 2, 4))
```

That `if` runs once, when the factory is called, and never again: the produced function does no branching at all. The choice of formula is paid for at construction time, and every later call to `bc0` or `bc1` pays only for the arithmetic. ggplot2 works the same way. `scale_color_brewer(palette = "Set1")` calls a factory, `pal_brewer()`, that returns a function mapping values to colors, so the palette is looked up once when the scale is built and the returned function only does the mapping. Separating the cost of *choosing* behavior from the cost of *running* it is called staged computation, and every factory you write does a version of it.

The next section starts from the other direction: instead of building a function from parameters, you take a function that already exists and modify its behavior.


### Exercises {.unnumbered}

1. Write a factory `between` that takes `low` and `high` and returns a function that keeps only elements of a vector that fall in the range `(low, high)`. Test it on `1:20` with bounds 5 and 15.
2. Write a factory `make_counter` that returns a function with no arguments. Each time the returned function is called, it should return the next integer (1, 2, 3, ...). Hint: use `<<-` to modify a variable in the enclosing environment.


## Memoization {#sec-memoization}

Sometimes you already have the right function, and the only problem is speed:

```{r}
#| message: false
library(memoise)

slow_square <- function(x) {
  Sys.sleep(1)
  x ^ 2
}

fast_square <- memoise(slow_square)
```

```{r}
system.time(fast_square(10))
system.time(fast_square(10))
```

The first call takes about a second. The second is instant: `memoise()` remembered the argument and returned the stored result without running the body. Wrapping a function so that it caches its results this way is *memoization*\index{memoization}. `forget()` clears the cache:

```{r}
forget(fast_square)
```

A cache only pays off when inputs recur: expensive computations called repeatedly with the same arguments, API calls, simulation steps, recursive algorithms like Fibonacci that revisit the same subproblems, parsing the same file twice. It goes wrong on functions with side effects (a memoized plot is drawn once and never again), on functions whose inputs are large or new every time (the cache grows without bound), and on functions that depend on external state such as database contents or the system clock, where the cached answer goes stale.


### Exercises {.unnumbered}

1. Write a function `slow_sum` that takes a vector, sleeps for 1 second, and returns the sum. Memoize it. Call it twice with the same input and verify the second call is fast.
2. What happens if you memoize `rnorm`? Try `memo_rnorm <- memoise(rnorm); memo_rnorm(5); memo_rnorm(5)`. Is this useful?


## Function operators {#sec-function-operators}

What do you do when a function might fail on some inputs, but you need to apply it to hundreds of values and cannot afford to let one error kill the whole pipeline? You could wrap every call in `tryCatch`. Or you could wrap the function itself once:

```{r}
#| message: false
library(purrr)

safe_log <- safely(log)
safe_log(10)
safe_log("a")
```

`safely()`\index{safely()} took `log` and returned a new function that never errors; it returns a list with `$result` and `$error` instead. A function that takes a function and returns a modified function is a *function operator*\index{function operator}\index{adverb}\index{error handling}: a factory whose raw material is a function rather than a number or a string.

`possibly()`\index{possibly()} is simpler: it returns a default value on error:

```{r}
careful_log <- possibly(log, otherwise = NA)
careful_log(10)
careful_log("a")
```

`quietly()`\index{quietly()} captures messages, warnings, and output as list components instead of printing them:

```{r}
quiet_log <- quietly(log)
quiet_log(-1)
```

You can write your own operators just as easily:

```{r}
with_logging <- function(f) {
  force(f)
  function(...) {
    cat("Calling function with", length(list(...)), "argument(s)\n")
    f(...)
  }
}

logged_mean <- with_logging(mean)
logged_mean(1:10)
```

`with_logging()` calls `force(f)` for the same reason `power()` did. A function operator is a factory whose input happens to be a function, so the lazy evaluation trap from @sec-lazy-eval-trap applies: without `force()`, `f` is a promise, and if the variable it points to changes before the operator's result is called, you get the wrong function.

::: {.callout-note collapse="true"}
## Decorators

Python calls this wrapping a *decorator* and gives it syntax: `@decorator` above a function definition replaces the function with the wrapped version. The pattern is older than the syntax; it is one of the twenty-three in the Gang of Four's *Design Patterns* (1994). R has no `@`, but `safely()`, `memoise()`, and `with_logging()` are all decorators: they change behavior by wrapping, leaving the original function untouched.
:::


### Exercises {.unnumbered}

1. Use `safely()` and `map()` to apply `log()` to the list `list(1, -1, "a", 10)`. Extract the results and the errors separately.
2. Write a function operator `with_timer` that wraps a function so it prints the elapsed time each time it is called. Test it with `Sys.sleep`.
3. What does `possibly(possibly(log, NA), NA)` do? Is double-wrapping useful?


## Composing factories and operators {#sec-composing-factories-operators}

Each of these patterns (factories, operators, functionals like `map()`) does one small thing, and they combine:

```{r}
safe_log <- safely(log)
results <- map(list(1, -1, "a", 10), safe_log)

str(results)
```

`safely(log)` is a function operator applied to `log`, and the result goes to `map()`. What if you need both safety *and* logging on the same function?

```{r}
safe_logged_log <- with_logging(safely(log))
safe_logged_log(10)
safe_logged_log("a")
```

The inner operator (`safely`) handles errors; the outer one (`with_logging`) adds logging. Each layer does one thing, and you combine them to get the behavior you want.

A factory builds specialized functions, an operator modifies them, and `map()` applies them across inputs.

```{r}
formatters <- map(c("$", "EUR ", "GBP "), make_formatter)

map(formatters, \(f) f(100))
```

`map()` over a vector of prefixes produces a list of formatters, then `map()` over the formatters applies each one. A list of functions is data you can iterate over, filter, compose, and pass around just like a list of numbers. @sec-reduce-and-accumulate uses that idea to collapse entire sequences into single values.
