# Reduce and accumulate {#sec-reduce-and-accumulate}

```{r}
#| include: false
source(here::here("_common.R"))
library(purrr)
library(dplyr)
```

You have a list of data frames from an API and you need to merge them all into one. `map()` will not help: it treats every element in isolation and never looks at what came before. What you need is a way to take two things, combine them, feed the result into the next combination, and keep going until one thing remains. That operation is `Reduce()`.

`map()` (@sec-iteration-without-loops) replaces loops that apply a function element by element. `Reduce()` replaces the other kind: loops that carry state from one pass to the next, like a running total or a chain of merges that folds data frames together one by one.


## Left fold {#sec-left-fold}

`Reduce()`\index{Reduce()} takes a binary function and a list, then combines the elements pairwise from left to right:

```{r}
Reduce(`+`, 1:5)
```

That computes `((((1 + 2) + 3) + 4) + 5)`. The first call is `f(1, 2)`, which gives 3; then `f(3, 3)`, giving 6; then `f(6, 4)`, giving 10; then `f(10, 5)`, giving 15. Each intermediate result becomes the left argument of the next call, which is why this is called a *left fold*\index{fold}\index{left fold} (Haskell's `foldl`): the parentheses nest to the left.

With a named function the folding structure becomes visible:

```{r}
Reduce(paste, c("the", "cat", "sat", "on", "the", "mat"))
```

Each step takes the accumulated string and pastes the next word onto it. The accumulator grows; the list shrinks. When nothing remains, the accumulator is the answer.


## The init argument {#sec-reduce-init}

What does `Reduce(f, x)` do when `x` has one element? It returns that element without calling `f`. And with zero elements?

```{r}
Reduce(`+`, list())
```

`NULL`, because there is nothing to combine and `Reduce()` has no idea what nothing adds up to. Anything downstream that expected a number now has `NULL` instead. The `init` argument tells it:

```{r}
Reduce(`+`, list(), init = 0)
```

With `init`, the fold starts by calling `f(init, x[[1]])` instead of `f(x[[1]], x[[2]])`, so it always has something to work with, even on an empty list. The value to pass is the one that leaves the operation unchanged: 0 for addition, 1 for multiplication, `""` for `paste0`, `NULL` for `c()`. @sec-no-scalars called this the identity element\index{identity element}. For `intersect` you would want the universal set, which R has no way to write, so you typically peel off the first element yourself.

::: callout-tip
## Opinion
Always pass `init` to `Reduce()`. Without it, your code returns `NULL` on empty input and behaves subtly differently on length-1 input. The identity element makes the fold total: it works for every list, including the empty one, and it still works six months from now when some upstream function returns zero results.
:::

::: {.callout-note collapse="true"}
## Folding a monoid

An associative\index{associativity} operation with an identity element is a *monoid*\index{monoid}, the structure from @sec-no-scalars that has since reappeared as verb composition, the pipe, and ggplot2 layers. `Reduce()` is the operation that collapses one: given a list and an associative binary function, it produces a single value. Piping is a fold over a list of functions with composition as the operation and the identity function as `init`; building a ggplot layer by layer is a fold over a list of layers with `+` as the operation and `ggplot()` as `init`. Associativity is also what makes a fold splittable: chunks can be folded separately and their results folded again without changing the answer, which is the shape of Google's MapReduce (Dean and Ghemawat, 2004), `map()` followed by a monoid fold across a warehouse of machines.
:::


## Right fold {#sec-right-fold}

`Reduce()` folds from the left by default. Setting `right = TRUE` reverses the nesting\index{right fold}:

```{r}
Reduce(`-`, 1:4)
Reduce(`-`, 1:4, right = TRUE)
```

The left fold computes `(((1 - 2) - 3) - 4) = -8`. The right fold computes `(1 - (2 - (3 - 4))) = -2`. For associative operations (addition, multiplication, `paste0`, `c()`, `intersect`, `union`) the direction does not matter; you get the same answer either way, and left fold is the natural default. The difference only surfaces for non-associative operations: subtraction, division, exponentiation.

In R, `right = TRUE` is rarely needed. If you find yourself reaching for a right fold, ask first whether your binary function simply has its arguments in the wrong order.

::: {.callout-note collapse="true"}
## Haskell's two folds

Haskell has `foldl` and `foldr` as separate functions, and there the choice matters. With lazy evaluation a right fold can stop early, even on an infinite list, when the combining function does not need its second argument; a left fold over a long list builds a tower of unevaluated thunks before anything is added. R's `Reduce()` evaluates eagerly, so `right = TRUE` only changes the nesting order.
:::

### Exercises {.unnumbered}

1. Use `Reduce()` to compute the product of the numbers 1 through 6 (i.e., 6 factorial). Check your answer against `factorial(6)`.
2. What is `Reduce(intersect, list(c(1,2,3,4), c(2,3,4,5), c(3,4,5,6)))`? Work through the steps by hand, then verify.
3. Write a `Reduce()` call that merges three data frames by a shared `"id"` column. (Hint: `merge` is a binary function.)
4. Explain why `Reduce(`/`, c(120, 2, 3, 4, 5))` gives 1 with a left fold and something different with a right fold.


## accumulate: keeping the intermediates {#sec-accumulate}

Sometimes you want every intermediate step, and only the last one comes out of `Reduce()`: the running total after each deposit, the growing sentence after each word, the high-water mark after each measurement. Base R handles this by adding a single argument:

```{r}
Reduce(`+`, 1:5, accumulate = TRUE)
```

Running totals: `1, 3, 6, 10, 15`. This is the *scan*\index{scan} operation (Haskell's `scanl`, APL's `\`). Where fold collapses a list to a single value, scan collapses it to a list of all the values the fold passed through on its way to the answer.

```{r}
Reduce(max, c(3, 1, 4, 1, 5, 9, 2, 6), accumulate = TRUE)
```

A running maximum, where each value is the largest seen so far. The 1s and the 2 cannot drag the maximum back down; it can only rise or hold steady.

```{r}
Reduce(paste, c("one", "two", "three", "four"), accumulate = TRUE)
```

A sentence assembling itself one word at a time.


## purrr::reduce() and purrr::accumulate() {#sec-purrr-reduce}

The purrr package provides `reduce()`\index{reduce()} and `accumulate()`\index{accumulate()} as tidyverse-flavored alternatives:

```{r}
reduce(1:5, `+`)
accumulate(1:5, `+`)
```

Same results. The differences from base `Reduce()` are syntactic: `.x` and `.f` instead of positional arguments, `.init` instead of `init`, and `.dir = "backward"` instead of `right = TRUE`. The purrr versions also accept lambda-style formulas, though anonymous functions with `\()` are clearer.

purrr also has `reduce2()`, which has no base equivalent: it folds over two lists in parallel, handing an extra argument from the second list at each step:

```{r}
reduce2(
  c("a", "b", "c"),
  c("-", "+"),
  \(acc, val, sep) paste0(acc, sep, val)
)
```

The binary function receives three arguments: the accumulator, the current element from the first list, and the corresponding element from the second list. The second list must be one element shorter than the first (or the same length if `.init` is provided), because the first element of the first list seeds the accumulator.


## Practical patterns {#sec-reduce-patterns}

When `split()` or repeated API calls hand you a list of data frames, `Reduce(rbind, dfs)` stacks them:

```{r}
dfs <- list(
  data.frame(x = 1:2, y = c("a", "b")),
  data.frame(x = 3:4, y = c("c", "d")),
  data.frame(x = 5:6, y = c("e", "f"))
)
Reduce(rbind, dfs)
```

Each step binds two frames and copies the growing result, so on a long list the cost is O(n^2^). `do.call(rbind, dfs)` binds all of them in one call, and `list_rbind()` from @sec-typed-variants does the same for tibbles.

Walking into a deeply nested structure one key at a time is a fold too:

```{r}
nested <- list(a = list(b = list(c = 42)))
Reduce(\(x, i) x[[i]], c("a", "b", "c"), init = nested)
```

Each step indexes one level deeper. The fold replaces a chain of `[[` calls. (purrr's `pluck()` does the same thing with better error messages.)

Set operations extend to any number of sets the same way:

```{r}
sets <- list(c(1,2,3,4), c(2,3,4,5), c(3,4,5,6))
Reduce(intersect, sets)
Reduce(union, sets)
```

And `accumulate()` can model any sequential process where each step depends on the previous state. A bank balance after each deposit and withdrawal:

```{r}
transactions <- c(100, -20, -30, 50, -10)
accumulate(transactions, `+`)
```

The running balance after each transaction. For more complex state, where the accumulator needs to carry multiple fields rather than a single number, use a list as the accumulator and a function that unpacks, updates, and repacks it.

### Exercises {.unnumbered}

1. Given `words <- c("R", "is", "a", "functional", "language")`, use `accumulate()` to produce the vector `c("R", "R is", "R is a", "R is a functional", "R is a functional language")`.
2. Write a `Reduce()` call that finds the intersection of the columns present in a list of data frames. (Hint: use `names()` and `intersect`.)
3. Use `accumulate()` to compute a running product of `c(2, 3, 5, 7)`. Verify that the last element equals `prod(c(2, 3, 5, 7))`.


## Row-wise folds with c_across() {#sec-c-across}

`across()` (@sec-across-revisited) applies a function down each column. Folding across the columns *within one row* is a different direction, and it starts with `rowwise()`, which tells dplyr to treat each row as a group of one:

```{r}
penguins <- palmerpenguins::penguins

penguins |>
  rowwise() |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  ungroup() |>
  head(3)
```

`c_across()` gathers the named columns of the current row into a vector, so any summary function can take it:

```{r}
df <- tibble(a = 1:3, b = 4:6, c = 7:9)
df |>
  rowwise() |>
  mutate(total = sum(c_across(a:c))) |>
  ungroup()
```

Each row's `a`, `b`, and `c` values are collected into a vector, summed, and stored in `total`: the same fold `Reduce()` does down a list, run across the columns of a row instead.

`rowwise()` + `c_across()` processes one row at a time in R, so it scales poorly. On 100,000 rows, `rowSums()` finishes in milliseconds because it is vectorized in C; the `rowwise()` version takes tens of seconds. Use `c_across()` when the operation is genuinely complex or when you need tidyselect helpers to pick columns. Otherwise use `rowSums()` or `pmax()`.

### Exercises {.unnumbered}

1. Using `rowwise()` and `c_across()`, compute the maximum of columns `a`, `b`, and `c` for each row of `df <- tibble(a = c(3,1,4), b = c(1,5,9), c = c(2,6,5))`.
2. Rewrite the `c_across()` sum example using `rowSums()` instead. Which is clearer? Which is faster? (Try `bench::mark()` on a larger data frame.)
3. Use `rowwise()` and `c_across(where(is.numeric))` to compute the range (max minus min) of every numeric column for each row of `penguins`. Why would this be hard to do without `c_across()`?


## When to fold and when not to {#sec-when-to-fold}

`Reduce()` is the right tool when you have a binary operation and a list of things to combine: merging data frames, chaining set operations, collapsing nested structures. It is the wrong tool when a vectorized function already exists. `sum(x)` is faster and clearer than `Reduce(`+`, x)`. `paste(x, collapse = " ")` beats `Reduce(paste, x)`. The base R cumulative functions (`cumsum`\index{cumsum()}, `cumprod`\index{cumprod()}, `cummax`, `cummin`) are compiled C and will always outperform `accumulate(`+`, x)`. Reach for `Reduce()` when your binary operation has no built-in vectorized cousin.

`Reduce()` also becomes awkward when the accumulator needs complex state. If your fold carries a list with multiple fields and each step updates several of them conditionally, you are writing a state machine and the fold is in the way. An explicit loop or a recursive function (@sec-recursion) shows the state updates where the reader can see them.

In 1999 Graham Hutton\index{Hutton, Graham} showed that fold can express *any* function on finite lists. That is a universality result, not practical advice. Just because something *can* be written as a fold does not mean it will be readable six months later.

::: callout-tip
## Opinion
Fold is the underused functional. Next time you write a loop that carries an accumulator variable, growing it with each pass, try `Reduce()` first. You might not need that loop at all.
:::
