# Contracts and defensive code {#sec-contracts-and-defensive-code}

```{r}
#| include: false
source(here::here("_common.R"))
```

You have written a function that computes a rate from two columns of a data frame. Someone passes it a data frame where one column is character instead of numeric. R does not crash; `mean()` on a character vector returns `NA` with a warning, and your function divides `NA` by a number and returns `NA`. The warning scrolls past in a long pipeline. The `NA` propagates through `mutate()`, `summarise()`, `left_join()`, and into a CSV that feeds a report. Three weeks later a collaborator asks why half the rates in the report are missing. You trace it back to a column that was read as character because one cell contained the string `"N/A"`.

Every line of code between the bad input and the discovery was wasted work. The function that produced the `NA` could have stopped the pipeline in the first line, with a message that said exactly what went wrong. It didn't, because nobody wrote that check. This was the problem a young man faced in the 1980s while working at Électricité de France, where a wrong number in a pipeline could mean a miscalculated load on a reactor.

After nine years of watching functions accept garbage and hand it downstream, he left and decided to build a programming language where that could not happen. His name was Bertrand Meyer\index{Meyer, Bertrand}; the language he was building was called Eiffel\index{Eiffel (language)}, which was defined by one central idea he called *Design by Contract*\index{Design by Contract}. It was simple: every function states a precondition\index{precondition} (what must be true before it runs) and a postcondition\index{postcondition} (what it guarantees after it runs). It also states an invariant\index{invariant}: what stays true throughout. If a precondition fails, the caller broke the deal. If a postcondition fails, the function broke it. Either way, execution stops with a message that names the broken obligation.

R has no contract\index{contract} syntax. But `stop()` at the top of a function is a precondition, `stopifnot()`\index{stopifnot()} before the return value is a postcondition, and the validator in a constructor/validator/helper pattern (@sec-s3-classes-properly) is a class invariant. The tools are manual; the discipline is the same. State your assumptions, halt the moment one is violated, and make the error message say what went wrong.


## The case for validation {#sec-case-for-validation}

R is dynamically typed, which means nothing stops you from passing a character vector to a function that expects a number. Without validation, the error surfaces deep inside the function body, far from the actual mistake:

```{r}
#| error: true
add_tax <- function(price, rate) {
  price * (1 + rate)
}

add_tax("42", 0.2)
```

The error message says `non-numeric argument to binary operator`. If you already know what went wrong, that is enough; if you don't, it is useless. The real problem is that someone passed `"42"` instead of `42`, but R doesn't tell you that. It tells you about a multiplication that failed, and leaves you to work out which argument was the string.

With validation, the error surfaces immediately:

```{r}
#| error: true
add_tax <- function(price, rate) {
  if (!is.numeric(price)) {
    stop(sprintf("price must be numeric, got %s", class(price)[1]))
  }
  price * (1 + rate)
}

add_tax("42", 0.2)
```

Check inputs at the boundary, where data enters your function, not in the guts where it is consumed. The principle is short: *fail fast, fail loud, fail clear*\index{defensive programming}.

R has no static types and no compile-time checks. Every precondition you want enforced must be enforced by code you write. That sounds like a burden, but it has an upside: every assumption is visible in the function body when you revisit it six months later. `if (!is.numeric(price)) stop(...)` is verbose, but it is also impossible to miss, and it forces the caller to deal with the problem before the bad value propagates through fifty lines of downstream code.

::: {.callout-note collapse="true"}
## Two missing checks
On 4 June 1996, 37 seconds into its first flight, the Ariane 5 rocket converted a 64-bit float holding its horizontal velocity into a 16-bit signed integer. Ariane 4 had flown for years with that conversion, and its velocities had always fit. Ariane 5 flew faster; the value overflowed, the guidance computer shut down, and the rocket and its payload, worth about 500 million euros, were destroyed. The conversion had been left unprotected on purpose, to keep the processor load down. A study of seven NASA spacecraft, published in 2004, found the same shape in a quieter setting: most safety-critical software changes after launch added handling for situations the original requirements had never mentioned. The requirements had rarely been wrong; they had been incomplete. When your function returns `NA` instead of stopping on bad input, it is in the same category.
:::

::: {.callout-note collapse="true"}
## The type theory connection
A `stop()`\index{assertion} that fires on wrong input is a runtime proof by contradiction: "if this value were numeric, execution would continue; it isn't, so we halt." The Curry-Howard correspondence formalizes this: types are propositions, programs are proofs, and a function with signature `A -> B` is a proof that "if A then B." Dependently typed languages (Idris, Agda) encode preconditions in the type system itself, making runtime `stop()` calls unnecessary. R sits at the opposite end of that spectrum. The question both approaches answer is the same: "what must hold for this computation to be sound?"
:::


## `stop()`, `warning()`, and `message()` {#sec-stop-warning-message}

R gives you three signalling functions, each pitched at a different severity.

`stop("msg")` halts execution and signals an error\index{error}. Use it when the function cannot produce a correct result.

```{r}
#| error: true
divide <- function(x, y) {
  if (y == 0) stop("y must not be zero")
  x / y
}

divide(10, 0)
```

`warning("msg")` continues execution but signals a warning\index{warning}, which makes it appropriate when the function can proceed but the result might surprise the caller.

```{r}
safe_log <- function(x) {
  if (any(x <= 0)) warning("non-positive values replaced with NA")
  x[x <= 0] <- NA
  log(x)
}

safe_log(c(1, -2, 3))
```

The result comes back, with the warning printed above it. Had the function called `log()` on the original vector, R would have added a second warning of its own, `NaNs produced`, and the caller would have seen two messages for one problem.

`message("msg")`\index{message} prints informational output: progress updates, status reports, diagnostics. `suppressMessages()` silences these, which is why `message()` is better than `cat()` for status output. It respects the caller's wishes.

```{r}
load_data <- function(path) {
  message(sprintf("Reading %s...", path))
  # ... actual loading code ...
}
```

The decision rule: if the result would be wrong, `stop()`. If the result is correct but surprising, `warning()`. If you are just providing status, `message()`. But where exactly is the line between "surprising" and "wrong"?

::: callout-tip
## Opinion
Never use `warning()` when you should use `stop()`. A warning that scrolls past unread is worse than no warning at all; it gives you the illusion of safety while the bad value keeps propagating. If you are not confident the function produced a correct result, don't return one.
:::


## Writing informative errors {#sec-informative-errors}

The difference between a helpful error and a frustrating one is the difference between ten seconds of debugging and ten minutes. Most of the gap comes down to one thing: does the message tell you what was expected and what was received?

```{r}
#| eval: false
# Bad
stop("invalid input")

# Good
stop(sprintf("x must be numeric, got %s", class(x)[1]))

# Better
stop(sprintf(
  "x must be a single positive number, got %s of length %d",
  class(x)[1], length(x)
))
```

Say what was *expected*, say what was *received*. The caller should be able to fix the problem without reading your source code.

`stopifnot()` is base R's assertion function, terse and good for internal checks where you trust the caller to interpret the output:

```{r}
#| error: true
normalize <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  x / sum(x)
}

normalize("abc")
```

The auto-generated error message is functional but cryptic. Since R 4.0, you can name the assertions to control the message:

```{r}
#| error: true
normalize <- function(x) {
  stopifnot(
    "x must be numeric" = is.numeric(x),
    "x must not be empty" = length(x) > 0
  )
  x / sum(x)
}

normalize("abc")
```

Named `stopifnot()` gives you the conciseness of assertions with messages you control. For functions that other people will call, this is the minimum standard. But what happens when you have ten arguments to validate, each with its own type, range, and length constraint?


### Exercises {.unnumbered}

1. The following function has a bad error message. Rewrite it so the error says what was expected and what was received.

    ```r
    compute_bmi <- function(weight_kg, height_m) {
      if (!is.numeric(weight_kg) || !is.numeric(height_m)) {
        stop("bad input")
      }
      weight_kg / height_m^2
    }
    ```

2. Rewrite the validation using named `stopifnot()`. Does the error message improve?

3. What error message does `stopifnot(is.numeric("hello"))` produce? What about `stopifnot("x must be numeric" = is.numeric("hello"))`?


## Common validation patterns {#sec-validation-patterns}

A function that expects a numeric scalar but receives a character vector of length three will fail somewhere. The question is where. This function checks everything it depends on before doing any work:

```{r}
#| error: true
top_n_share <- function(x, n) {
  if (is.null(x)) stop("x must not be NULL")
  if (!is.numeric(x)) stop(sprintf("x must be numeric, got %s", class(x)[1]))
  if (anyNA(x)) stop("x must not contain NA values")
  if (length(n) != 1 || n < 1) stop("n must be a single positive number")
  sum(sort(x, decreasing = TRUE)[seq_len(n)]) / sum(x)
}

top_n_share(c(5, 3, 2), 2)
top_n_share(c(5, NA, 2), 2)
```

The same four kinds of check come up in almost every function: the type (`is.numeric()`, `is.character()`, `is.data.frame()`), the length (`length(x) == 1` for a scalar, `length(x) > 0` for a non-empty input), the range (`x > 0`, or `x >= 0 && x <= 1` for a probability), and the two absences, `is.null(x)` and `anyNA(x)`. A function that reads a file adds `file.exists(path)` before it opens anything:

```{r}
#| eval: false
if (!file.exists(path)) stop(sprintf("file not found: %s", path))
```

One more check comes up so often that base R has a function for it. When an argument is a string chosen from a fixed set of options, list the options in the signature and let `match.arg()` compare against them:

```{r}
my_summary <- function(x, method = c("mean", "median", "trimmed")) {
  method <- match.arg(method)
  switch(method,
    mean    = mean(x),
    median  = median(x),
    trimmed = mean(x, trim = 0.1)
  )
}

my_summary(1:10)
my_summary(1:10, "median")
my_summary(1:10, "tri")
```

```{r}
#| error: true
my_summary(1:10, "xyz")
```

`match.arg()` did three things. With no `method` supplied it picked the first option, so `my_summary(1:10)` used `"mean"`. It accepted `"tri"` as a partial match for `"trimmed"`. And it rejected `"xyz"` with a message that lists the valid options, which is why it is preferred over a hand-written `%in%` check: the choices live in the function signature, where `?my_summary` shows them, and the error repeats them. But all of these patterns share a problem: writing them out by hand, over and over, for every function in a package.


### Exercises {.unnumbered}

1. Write a function `rescale()` that takes a numeric vector `x` and a `method` argument accepting `"minmax"` or `"zscore"`. Use `match.arg()` for validation. Test that partial matching works.

2. Add validation to the following function using `stopifnot()` or `if`/`stop()`. Check that `data` is a data frame, `n` is a single positive integer, and `col` is a string that exists as a column name.

    ```r
    top_rows <- function(data, col, n = 5) {
      data[order(data[[col]], decreasing = TRUE), ][seq_len(n), ]
    }
    ```


## The checkmate package {#sec-checkmate}

Writing the same `if (!is.numeric(x)) stop(...)` pattern for the fifteenth time in a package gets old fast, and each repetition is another chance to misspell a class name or forget a length check. The `checkmate` package provides concise, fast assertion functions that handle the boilerplate:

```{r}
#| eval: false
library(checkmate)

process <- function(data, method = "fast", threshold = 0.5) {
  assert_data_frame(data, min.rows = 1)
  assert_choice(method, choices = c("fast", "accurate"))
  assert_number(threshold, lower = 0, upper = 1)
  # ...
}
```

Each `assert_*` function throws an informative error if the check fails. No `if`/`stop` boilerplate, no `sprintf` formatting. The function names are self-documenting: `assert_numeric`, `assert_character`, `assert_file_exists`, `assert_flag` (for single TRUE/FALSE), and dozens more.

`checkmate` also provides two softer variants. `test_*` functions return `TRUE` or `FALSE` for use in conditional logic; `check_*` functions return the error message as a string (or `TRUE` on success) for custom handling.

```{r}
#| eval: false
# Conditional logic
if (test_numeric(x)) {
  mean(x)
} else {
  length(x)
}

# Custom error message
msg <- check_numeric(x, lower = 0)
if (!isTRUE(msg)) stop("Custom prefix: ", msg)
```

::: callout-tip
## Opinion
For package code with many validated arguments, `checkmate` pays for itself in readability and consistency. For scripts and one-off functions, `stopifnot()` and `match.arg()` are enough. Don't add a dependency for two assertions.
:::


## `tryCatch()` and error handling {#sec-trycatch}

Everything so far has been about signalling your own errors. But sometimes the error comes from somewhere else: a file that doesn't exist, a web API that times out, a parsing function that chokes on malformed input. You need to *catch* those errors and decide what to do about them.

`tryCatch()`\index{tryCatch()} catches an error, warning, or message and runs a handler:

```{r}
result <- tryCatch(
  log("a"),
  error = function(e) NA
)

result
```

Instead of crashing, `log("a")` returns `NA`. The error handler receives the condition object `e` and returns a fallback value. This is *recovery*: you know an error might happen, and you have a plan for it.

`try()` is a simpler wrapper that returns the result on success, or an object of class `"try-error"` on failure:

```{r}
result <- try(log("a"), silent = TRUE)
class(result)
```

`try()` is convenient for quick scripts, but `tryCatch()` lets you run a handler and catch warnings and messages, not just errors.

For functional programming, `purrr::safely()` wraps a function to make it error-tolerant. It returns a list with `$result` and `$error`, designed for use with `map()` when applying a function that might fail on some elements:

```{r}
safe_log <- purrr::safely(log)
safe_log(10)
safe_log("a")
```

`purrr::possibly()` is a variation: it returns a default value on failure instead of the `$result`/`$error` list. Errors, warnings and messages are all *conditions*\index{condition}, and `tryCatch()` can take a handler for each.

::: {.callout-note collapse="true"}
## Let it crash
In the mid-1980s a team at Ericsson was looking for a language to program telephone switches, machines expected to run for years without a restart. The language they built, Erlang, takes the opposite approach to this chapter: instead of validating inputs so that a process never fails, it lets processes crash and assigns a *supervisor* process to notice the crash and start a clean copy. Ericsson has claimed nine nines of availability, 99.9999999%, for one of its Erlang-based switches, though the figure comes from a customer's report on a single installation rather than a systematic measurement. `tryCatch()` is a mild version of the same idea: the error is expected, and there is a plan for it.
:::

::: {.callout-note collapse="true"}
## Either and Result
`tryCatch(expr, error = function(e) NA)` produces one of two shapes: the value of `expr`, or whatever the handler returns. Typed functional languages give that pair a name. Haskell's `Either` holds either an error or a result, Rust's `Result` does the same, and `purrr::safely()` is the R version: a list with a `$result` slot and an `$error` slot, exactly one of them filled.
:::

::: callout-tip
## Opinion
Use `tryCatch()` when you have a recovery plan. Don't use it to silently swallow errors. An error you hide is a bug you invite.
:::


### Exercises {.unnumbered}

1. Write a function `safe_read` that takes a file path and returns the file contents (via `readLines()`), or `NULL` if the file doesn't exist. Use `tryCatch()`.

2. Use `purrr::map()` and `purrr::possibly()` to apply `as.numeric()` to the list `list("1", "two", "3", "four")`. Use `NA` as the default.


## `rlang::abort()` and cli {#sec-rlang-abort}

`rlang::abort()` is a modern replacement for `stop()` that automatically saves a backtrace (the chain of function calls that led to the error) and supports *condition classes* for selective catching:

```{r}
#| eval: false
rlang::abort(
  message = "x must be positive",
  class = "validation_error",
  x = x
)
```

A caller can now catch `validation_error` specifically, while letting other errors propagate:

```{r}
#| eval: false
tryCatch(
  my_function(x),
  validation_error = function(e) {
    message("Validation failed: ", e$message)
  }
)
```

The `cli` package takes this further. `cli::cli_abort()` produces structured, formatted error messages with bullets and inline markup:

```{r}
#| eval: false
cli::cli_abort(c(
  "x" = "{.arg x} must be a positive number.",
  "i" = "You supplied {.cls {class(x)}} of length {length(x)}."
))
```

The `"x"` prefix marks the main error; `"i"` marks informational context. `{.arg x}` and `{.cls ...}` are cli's inline formatting: they render the argument name and class with special styling in terminals that support it.

If you have ever wondered why tidyverse error messages look so polished, with colored bullets and formatted argument names, `cli::cli_abort()` is the answer. You don't need cli-formatted errors in scripts, but knowing they exist helps you read the error messages that tidyverse packages produce.

`rlang` also provides `rlang::arg_match()`, a stricter alternative to `match.arg()`. It produces tidyverse-styled error messages and does *not* support partial matching, which makes it safer for package APIs where partial matching can cause subtle bugs.


### Exercises {.unnumbered}

1. Rewrite one of your `stop()` calls from the exercises above using `rlang::abort()` with a custom class (e.g., `"input_error"`). Then write a `tryCatch()` that catches only that class.
2. Use `cli::cli_abort()` to produce an error message with a bullet point showing the expected type and the actual type. Compare the output with a plain `stop()` message.


## Designing function interfaces {#sec-function-interface-design}

Validation is downstream of design. A well-designed function is hard to misuse in the first place, which means much of your defensive code never needs to exist.

Two ways of accepting a plot type:

```{r}
#| eval: false
my_plot <- function(data, type = c("scatter", "line", "bar")) {
  type <- match.arg(type)
  # ...
}

my_plot <- function(data, type = "scatter") {
  if (!type %in% c("scatter", "line", "bar")) stop("invalid type")
  # ...
}
```

Both reject `"pie"`. The first version also tells the caller what the options are, in the signature and in the help page, before anything goes wrong. In the second the choices are hidden inside the body, and the caller has to read the source or trigger the error to discover them.

The same logic applies to defaults. If a function needs a file path to work, `path` gets no default; giving it `NULL` and checking for `NULL` inside the body is validation code that a required argument would have made unnecessary. Optional arguments get sensible defaults. The data argument goes first, so the function works in a pipe (`data |> my_func()`), and the modifiers come after.

Use `...` sparingly. If you use it, document what it accepts and where it gets forwarded, because a misspelled argument name disappears into `...` without an error, and a function that quietly drops `na.rm = TRUE` is one of R's most common sources of invisible bugs.

An argument `x` that can be a vector, a data frame, or a formula is hard to validate and hard to document, because each accepted type needs its own branch. If you find yourself writing `if (is.data.frame(x)) ... else if (is.numeric(x)) ...`, consider splitting the function or using S3 dispatch.

This connects to @sec-s3-and-s7: S7 properties with type declarations are validation built into the class definition. When you declare that a property must be a positive numeric scalar, invalid objects cannot exist. Validation at construction time, enforced by the type system, is the strongest form of defensive programming.

::: callout-tip
## Opinion
The best validation is the kind you don't have to write. A function with a clear signature, named options via `match.arg()`, and typed S7 properties catches most mistakes before any `if`/`stop` check runs. Design the interface right, and defensive code becomes a safety net instead of a load-bearing wall.
:::


## Summary

Silent failures are the worst kind of bug: the function returns an answer, and the answer is wrong, and nothing tells you. Validate at the boundary. Write error messages that say what was expected and what was received. Catch errors only when you have a recovery plan. And design interfaces that make misuse difficult in the first place, so the validation you write covers less ground.

