# R as mathematics {#sec-r-as-mathematics}

```{r}
#| include: false
source(here::here("_common.R"))
```

Take R and remove everything except `function()` and the ability to call functions. No numbers. No booleans. No `if`. No `list()`. No `<-` for naming recursive calls. From those two things alone, try to build integers, conditionals, pairs, lists, and recursion.

Surprisingly, it works. Every one of those things can be reconstructed from `function()` alone. The `double <- function(x) x * 2` from @sec-first-function, the recursive descents in @sec-recursion, the closures that captured variables and carried them somewhere new, all of them are expressions in a formal system that predates computers by decades.

This chapter runs the experiment.


## Lambda calculus in one page {#sec-lambda-calculus}

Two expressions to start with:

```{r}
(function(x) x + 1)(5)
```

```{r}
(function(x) function(y) x + y)(3)(4)
```

The first builds a function and applies it in the same breath. The second builds a function that returns a function, applies it to 3, and applies the result to 4: a two-argument function rewritten as a chain of one-argument functions, the currying from @sec-arguments-defaults. Both use exactly three things: a variable, `function()` to build an abstraction over it, and a call to apply one. Those three are the entire lambda calculus\index{lambda calculus!in R}, the system Alonzo Church published in 1936 and that @sec-two-models introduced. Church wrote `λx. body` where R writes `function(x) body`, and `(f a)` where R writes `f(a)`:

| Lambda calculus | R |
|---|---|
| `λx. x` | `function(x) x` |
| `λx. x + 1` | `function(x) x + 1` |
| `(λx. x + 1)(5)` | `(function(x) x + 1)(5)` |
| `λx. λy. x + y` | `function(x) function(y) x + y` |

No numbers, no booleans, no `if`, no loops, and Church showed that the three are enough to express any computation, the same class of functions Turing's machine computes (@sec-computability).

The translation works because of a decision about scoping.\index{beta reduction!in R} When the inner `function(y) x + y` runs, it looks up `x` in the environment where it was *created*, not where it was called, which is what the lambda calculus requires: a variable refers to the nearest enclosing abstraction that binds it. R has had that rule, and unrestricted anonymous functions, since its first release (@sec-what-r-inherited).

So what can you build on that foundation, using nothing else?


### Exercises {.unnumbered}

1. Write the lambda calculus expression `λf. λx. f(f(x))` in R and apply it to `sqrt` and `256`. What do you get?
2. Write a curried `power` function: `power <- function(n) function(x) x^n`. Create `square <- power(2)` and `cube <- power(3)`. Test them.
3. What happens if you call `(function(x) function(y) x)(1)(2)`? Why? Think about which variable the inner function captures.


## Church booleans {#sec-church-booleans}

Suppose all you have is `function()`. How do you represent `TRUE` and `FALSE`? Start with two curried functions that differ in one letter:

```{r}
ch_true  <- function(a) function(b) a
ch_false <- function(a) function(b) b
```

```{r}
ch_true("yes")("no")
ch_false("yes")("no")
```

`ch_true` takes `a`, returns a function that takes `b` and ignores it, and hands back `a`. `ch_false` ignores the first argument and returns the second. That is Church's answer: a boolean\index{Church boolean} is a function that *chooses*. Give it two things and `TRUE` picks the first, `FALSE` the second, which is already `if/else`. Call a Church boolean with the "then" branch and the "else" branch and it selects one, with no dispatch mechanism, no keyword, and no syntax.

```{r}
ch_if <- function(cond) function(then_val) function(else_val) {
  cond(then_val)(else_val)
}

ch_if(ch_true)("heads")("tails")
ch_if(ch_false)("heads")("tails")
```

`ch_if` does almost nothing; it applies the boolean to the two branches, and the boolean does the work.

Now the logical connectives:

```{r}
ch_and <- function(p) function(q) p(q)(p)
ch_or  <- function(p) function(q) p(p)(q)
ch_not <- function(p) function(a) function(b) p(b)(a)
```

Watch `ch_and(p)(q)`: if `p` is true, it picks its first argument, which is `q` (so the result depends on whether `q` is also true); if `p` is false, it picks its second argument, which is `p` itself (false). The logic lives in the selection behaviour of the booleans themselves.

```{r}
# TRUE AND FALSE -> FALSE
ch_and(ch_true)(ch_false)("yes")("no")

# TRUE OR FALSE -> TRUE
ch_or(ch_true)(ch_false)("yes")("no")

# NOT TRUE -> FALSE
ch_not(ch_true)("yes")("no")
```

::: callout-tip
## Opinion

Church booleans make a quiet point about R: `if/else` is redundant. Conditional logic can be built from functions alone, because R's built-in `if` is syntactic sugar for something that `function()` can already do. A language whose core primitive is powerful enough to subsume its own control flow has a different character from one that added lambdas later.
:::


### Exercises {.unnumbered}

1. Implement `ch_xor` (exclusive or) using only Church booleans. Test it on all four input combinations.
2. Implement `ch_eq` that checks whether two Church booleans are equal. (Hint: `xor` followed by `not` works, or find a direct encoding.)
3. Write a function `to_r_bool` that converts a Church boolean back to R's `TRUE`/`FALSE`.


## Church numerals {#sec-church-numerals}

Numbers are harder to believe. Here are four of them:

```{r}
zero  <- function(f) function(x) x
one   <- function(f) function(x) f(x)
two   <- function(f) function(x) f(f(x))
three <- function(f) function(x) f(f(f(x)))
```

Each takes a function `f` and returns `f` composed with itself some number of times: `zero` never applies it, `three` applies it three times. To see what one of them means, pass it a concrete function and a starting value:

```{r}
three(function(x) x + 1)(0)
```

Three applications of "add one" to zero gives three. That is the conversion from Church numeral\index{Church numeral} to R integer:

```{r}
to_int <- function(n) n(function(x) x + 1)(0)

to_int(zero)
to_int(two)
to_int(three)
```

The successor function adds one more application of `f`:

```{r}
succ <- function(n) function(f) function(x) f(n(f)(x))
```

Read it this way: `succ(n)` returns a new numeral that, given `f` and `x`, first applies `f` `n` times (that is `n(f)(x)`), then applies `f` one more time on top.

```{r}
four <- succ(three)
to_int(four)
```

Build higher numbers:

```{r}
five <- succ(four)
six  <- succ(five)
to_int(six)
```

Addition could be `succ` repeated, but there is a direct form:

```{r}
add <- function(m) function(n) function(f) function(x) m(f)(n(f)(x))

to_int(add(two)(three))
```

Read `m(f)(n(f)(x))`: apply `f` to `x` `n` times, then stack `m` more applications of `f` on top, giving `m + n` in total.

Multiplication: `mult(m)(n)` means "apply `f` `n` times, repeated `m` times." Composing `n` copies of `f` gives a function that applies `f` `n` times; applying that composition `m` times gives `m * n`.

```{r}
mult <- function(m) function(n) function(f) m(n(f))

to_int(mult(two)(three))
to_int(mult(three)(three))
```

Exponentiation falls out naturally:

```{r}
power <- function(m) function(n) n(m)

to_int(power(two)(three))   # 2^3 = 8
to_int(power(three)(two))   # 3^2 = 9
```

`power(m)(n)` applies `m`, itself a function-repeater, `n` times: two applied three times is `2^3`. The operation that felt most abstract has the shortest definition, because applying a repeater repeatedly is what exponentiation means. Building numbers up is only half the story, though. Can we ask whether a numeral is zero?

The test is short:

```{r}
is_zero <- function(n) n(function(x) ch_false)(ch_true)

is_zero(zero)("yes")("no")
is_zero(three)("yes")("no")
```

`zero` ignores `f` and returns `x`, so it returns `ch_true`. Any nonzero numeral applies `f` at least once, and since `f` always returns `ch_false`, a single application replaces `ch_true` with `ch_false`.

We can also convert R integers to Church numerals. A `for` loop that reassigns one variable would fall into the trap from @sec-lazy-eval-trap, because `succ(n)` keeps `n` as a promise and forces it only later, by which time the variable holds the numeral that contains it. Recursion gives each `succ` its own `n`:

```{r}
from_int <- function(n) if (n == 0) zero else succ(from_int(n - 1))

to_int(from_int(5))
to_int(add(from_int(3))(from_int(4)))
```


### Exercises {.unnumbered}

1. Compute `mult(from_int(4))(from_int(5))` and convert back to an integer.
2. Compute `power(from_int(2))(from_int(9))` and convert it back. Then try `from_int(10)` in place of `from_int(9)`. What happens, and why? (`?options`, under `expressions`, is the place to look.)
3. The predecessor function (subtracting one) is famously tricky in Church encoding. Look it up, implement it in R, and test it. This was hard enough that Church himself struggled with it; his student Kleene figured it out, reportedly while getting anesthesia at the dentist.


## Church pairs and lists {#sec-church-pairs}

A pair\index{pair!Church encoding} is a container that holds two values. In Church encoding, a pair is a function that takes a selector and applies it to both values:

```{r}
pair <- function(a) function(b) function(f) f(a)(b)
fst  <- function(p) p(function(a) function(b) a)
snd  <- function(p) p(function(a) function(b) b)
```

`pair(a)(b)` stores `a` and `b` inside a closure, and `fst` and `snd` extract them by passing the appropriate selector (which, if you look closely, is just `ch_true` and `ch_false` again):

```{r}
p <- pair("hello")("world")
fst(p)
snd(p)
```

From pairs, you can build linked lists\index{linked list!Church encoding}. A list is either empty (nil) or a pair of a head and a tail:

```{r}
nil     <- function(f) function(x) x   # same shape as zero
is_nil  <- function(l) l(function(h) function(t) function(x) ch_false)(ch_true)
cons    <- function(h) function(t) pair(h)(t)
head_of <- function(l) fst(l)
tail_of <- function(l) snd(l)
```

`is_nil` passes a selector that ignores both arguments and returns `ch_false`, then passes `ch_true` as the default. If `l` is `nil`, it behaves like `zero`: it ignores the selector and returns the default. If `l` is a `cons` cell, which is a pair, the selector runs on head and tail, ignores both, and returns `ch_false`. The structure of the data does the dispatching.

Build a list and walk it:

```{r}
my_list <- cons("a")(cons("b")(cons("c")(nil)))

head_of(my_list)
head_of(tail_of(my_list))
head_of(tail_of(tail_of(my_list)))
```

Three nested function calls, and you have a linked list. No vectors, no `list()`, no memory allocation beyond closures. But one thing is still missing: none of these structures can refer to themselves.


### Exercises {.unnumbered}

1. Write a `length_of` function that counts the elements in a Church-encoded list. Use `to_int` and Church numerals, or use R's integers for simplicity.
2. Write `map_list` that applies a function to every element of a Church list and returns a new Church list.
3. Implement `reverse_list`. (Hint: fold from the left, consing onto an accumulator that starts as `nil`.)


## The Z combinator {#sec-z-combinator}

A recursive function refers to itself by name, but the lambda calculus has no names and no `<-`. @sec-fixed-point-combinators built the way around that: a template that takes the function to recurse on as an argument, and a *fixed-point combinator*\index{fixed-point combinator} that feeds the template to itself. The Y combinator\index{Y combinator} there ran in R only because R hands arguments over as promises. The Z combinator\index{Z combinator} makes the delay explicit by wrapping each self-application in a function of one argument, so it runs in strict languages too, and it is the version this chapter uses:

```{r}
Z <- function(f) {
  (function(x) f(function(v) x(x)(v)))(
    function(x) f(function(v) x(x)(v))
  )
}
```

`Z(f)` builds a function `x` that, when called, applies `f` to a delayed version of the recursive call. `function(v) x(x)(v)` means "do not call `x(x)` yet; wait until `v` is provided."

Use it to define factorial without ever naming the function:

```{r}
fact <- Z(function(self) function(n) {
  if (n == 0) 1 else n * self(n - 1)
})

fact(5)
fact(10)
```

`self` is not a name the function gives itself. It is a parameter, injected by `Z`. The function never refers to `fact`; it only refers to `self`, which `Z` arranges to be the function itself.

Fibonacci:

```{r}
fib <- Z(function(self) function(n) {
  if (n <= 1) n else self(n - 1) + self(n - 2)
})

sapply(0:10, fib)
```

A recursive list-sum, using nothing but Church-encoded lists and the Z combinator:

```{r}
church_sum <- Z(function(self) function(lst) {
  is_nil(lst)(0)(head_of(lst) + self(tail_of(lst)))
})

nums <- cons(10)(cons(20)(cons(30)(nil)))
church_sum(nums)
```

`is_nil(lst)` returns a Church boolean, and the boolean picks one of its two arguments. On the empty list it picks `0`, and the recursive branch, `head_of(lst) + self(tail_of(lst))`, is never evaluated, because it arrived as a promise (@sec-promises) and nothing forced it. That is the only reason the recursion stops.

::: {.callout-note collapse="true"}
## In a strict language

A language that evaluates every argument before the call would compute both branches, and the recursive one would run on the empty list, forever. The fix is the same one that turns Y into Z: wrap each branch in a function that takes a dummy argument, let the boolean choose a wrapper, and call the winner.

```{r}
church_sum_strict <- Z(function(self) function(lst) {
  is_nil(lst)(
    function(dummy) 0
  )(
    function(dummy) head_of(lst) + self(tail_of(lst))
  )("go")
})

church_sum_strict(nums)
```

The `"go"` argument is unused; it is the trigger that makes the chosen wrapper run.
:::

No `for`, no `while`, no `Recall()`, no named recursion. Just `function()`, application, and the Z combinator.

::: callout-tip
## Opinion

The Z combinator is not practical R code. Its value is conceptual: it proves that recursion is not a primitive operation. Self-reference can be *derived* from anonymous functions and application alone. That R can express this in code you can paste into a console and run is a direct consequence of its Scheme heritage.
:::


### Exercises {.unnumbered}

1. Define a recursive `length` function using `Z` that counts the elements of a Church list (without converting to R types).
2. Use `Z` to write a recursive `map` over Church lists.
3. The Y combinator from @sec-fixed-point-combinators also runs in R. Rewrite `church_sum` with `Y` in place of `Z`. Then change `Z` so that it forces the self-application before calling `f` (`function(x) { y <- x(x); f(y) }`) and try again. Which version loops forever, and why?


## The punchline {#sec-the-punchline}

Here is the inventory. Starting from nothing but `function()` and function application\index{Church encoding}:

- Booleans: `ch_true`, `ch_false`, `ch_and`, `ch_or`, `ch_not`
- Conditionals: `ch_if` (which turned out to be trivial, because booleans *are* conditionals)
- Natural numbers: `zero`, `succ`, and arithmetic (`add`, `mult`, `power`)
- Pairs and lists: `pair`, `fst`, `snd`, `cons`, `nil`
- Recursion: the Z combinator

No R integers, no R booleans, no `if`, no `list()`, no named recursive functions. Every one of those features reconstructed from `function()` alone.

This is Church's result from 1936. When @sec-two-models said that R descends from Church's model of computation, this is what that means concretely: `function()` is a universal computational primitive, and everything else is convenience built on top. The requirements are specific: first-class functions, closures, and anonymous functions with no restriction on their bodies. R has all three from Scheme. A language that restricts any of them (Python's one-expression `lambda`, C's function pointers without closures) can run only part of this chapter.

The connection between R and mathematics extends further, into the type system itself.


## Types as propositions {#sec-curry-howard}

In @sec-first-function a logician noticed that the rules for combining function types looked like the rules for combining propositions. Decades later, in 1969, William Howard wrote the correspondence out in full for constructive logic and typed lambda calculus\index{Curry-Howard correspondence}:

| Logic | Programming |
|---|---|
| Proposition | Type |
| Proof | Program |
| A implies B | Function type A -> B |
| A and B | Pair type (A, B) |
| A or B | Sum type (either A or B) |
| True | Unit type (a type with one value) |
| False | Empty type (a type with no values) |

When you write a function with signature `A -> B`, you are constructing a proof that if you have an `A`, you can produce a `B`. A function `integer -> character` proves that integers can be converted to characters. If no function of type `A -> B` can be written, the proposition "A implies B" has no proof.

R's type system is too loose to enforce this correspondence strictly (every R function can error, which is like a proof that cheats). Haskell, Agda, and Coq take it seriously enough to use programs as machine-checked proofs of theorems. But the connection is there in R: every function you write is, in a formal sense, a small proof.

Under this reading, named the Curry-Howard correspondence for the two people who saw it, a type is a proposition about which values are possible and a function is evidence that a transformation exists. What happens when you start looking at the *structure* of those propositions and transformations?


## Functors, monoids, and friends {#sec-category-connections}

Two of the patterns in this book already carry names from category theory\index{category theory}: the functor from @sec-purrr-map and the monoid from @sec-no-scalars. This section adds the laws that make them trustworthy, and a third pattern that connects them.

Run `lapply()` over a list:

```{r}
x <- list(1, 4, 9, 16)
lapply(x, sqrt)
```

`lapply()` applied `sqrt` to each element and gave back a list of the same length. The container's shape did not change; only the values inside it did. Strictly, the functor\index{functor} is the list construction itself: it sends a type `T` to "list of `T`" and a function `f` to "apply `f` to each element", and `lapply()` (and purrr's `map()`) is the part that lifts the function into the container.

What makes a functor trustworthy is that it obeys two laws. The *identity law* says mapping a do-nothing function changes nothing:

```{r}
identical(lapply(x, identity), x)
```

The *composition law* says mapping two functions in sequence gives the same result as mapping their composition in one step:

```{r}
identical(
  lapply(lapply(x, sqrt), log),    # map sqrt, then map log
  lapply(x, \(v) log(sqrt(v)))     # map the composition directly
)
```

A function that mapped elements but scrambled their relationships would break pipelines; chains of `lapply()` work reliably because both laws hold.

A monoid\index{monoid!category theory} is a set with an associative operation and an identity element: addition with 0, string concatenation with `""`. `Reduce()` (@sec-reduce-and-accumulate) is the operation that collapses one:

```{r}
Reduce(paste0, c("a", "b", "c"))
Reduce(`+`, 1:5)
```

It needs associativity so that the result does not depend on how the intermediate steps are grouped, and an identity element for the empty case. Subtraction fails the first requirement:

```{r}
(5 - 3) - 1
5 - (3 - 1)
```

Different answers. The operations that fold cleanly are the monoids, and when `Reduce()` surprises you, check whether the operation is one; if it is not, the folding direction matters (@sec-right-fold).

The third pattern connects two functors. `as.list()` turns an atomic vector into a list. Map a function before or after the conversion:

```{r}
x <- c(1, 4, 9)

# map then convert
a <- as.list(sqrt(x))

# convert then map
b <- lapply(as.list(x), sqrt)

identical(a, b)
```

The order does not matter. A conversion between containers that commutes with mapping in this way is a *natural transformation*, and the commuting itself is the naturality condition: `as.list()` changes the container without disturbing the relationship between the function and the data inside.

`pivot_longer()` and `pivot_wider()` (@sec-pivot-longer) go further: they are invertible, which makes them natural *isomorphisms*. Reversibility is a stronger property than naturality alone; it means no information is lost in the reshaping. You can pivot long and then pivot wide and arrive back where you started, because the transformation preserves everything.

None of this requires learning category theory. But these patterns keep showing up because they reflect mathematical structure, and R's functional design makes them visible.


### Exercises {.unnumbered}

1. `paste0` with `Reduce`: `Reduce(paste0, c("a", "b", "c"))` gives `"abc"`. What is the identity element for `paste0` as a monoid? Verify by adding it as the `init` argument: `Reduce(paste0, character(0), accumulate = FALSE)`.
2. Why is subtraction not a monoid? (Hint: check associativity.) What does this imply about using `Reduce` with `-`?
3. `lapply(list(1:3, 4:6, 7:9), rev)` applies `rev` inside each list element. Verify both functor laws (identity and composition) for this example.


## Practical mathematics in R {#sec-practical-math}

Church encodings prove R can do mathematics from nothing. But can it do mathematics that matters? Symbolic algebra, arbitrary precision, number theory: each requires capabilities that most R users never discover, because the packages that provide them sit outside the usual statistical workflow.

The `Ryacas` package connects R to the Yacas computer algebra system, which works with expressions rather than numbers:

```{r}
#| eval: false
library(Ryacas)

# Symbolic differentiation
yac_str("D(x) x^3 + 2*x")  # "3*x^2+2"

# Symbolic integration
yac_str("Integrate(x) x^2")  # "x^3/3"

# Simplification
yac_str("Simplify((x^2-1)/(x-1))")  # "x+1"
```

Where R's floating-point arithmetic gives you `0.1 + 0.2 != 0.3` (@sec-floating-point-traps), symbolic computation works with exact representations. There is no rounding because there are no decimals; expressions stay symbolic until you ask for a numerical result.

The `gmp` package provides big integers and big rationals:

```{r}
#| eval: false
library(gmp)

# Big integers
factorial(as.bigz(100))  # all 158 digits, exact

# Big rationals: 1/3 stays 1/3, no floating-point error
as.bigq(1, 3) + as.bigq(1, 6)  # 1/2, exactly
```

The `numbers` package has prime factorization, GCD, modular arithmetic, and related tools:

```{r}
#| eval: false
library(numbers)

primeFactors(2310)  # 2, 3, 5, 7, 11
GCD(48, 36)         # 12
isPrime(104729)     # TRUE
```

If you want to sharpen your R skills on mathematical puzzles, [Project Euler](https://projecteuler.net) is a good source. The first few problems are approachable with base R:

```{r}
# Euler problem 1: sum of multiples of 3 or 5 below 1000
x <- 1:999
sum(x[x %% 3 == 0 | x %% 5 == 0])
```

```{r}
# Euler problem 6: difference between sum-of-squares and square-of-sum
n <- 1:100
sum(n)^2 - sum(n^2)
```

Both solutions are one-liners because R's vectorized operations (@sec-vectorized-ops) make arithmetic on sequences natural.


### Exercises {.unnumbered}

1. Solve Project Euler problem 2: find the sum of even Fibonacci numbers below four million. (Hint: generate Fibonacci numbers with a while loop or `Reduce`, filter, sum.)
2. Install `gmp` and compute `factorial(as.bigz(200))`. How many digits does the result have? (Use `nchar(as.character(...))`.)
3. `1/3 + 1/3 + 1/3 == 1` returns `TRUE` in R, but `0.1 + 0.1 + 0.1 == 0.3` returns `FALSE`. Why? (Review @sec-floating-point-traps if needed.)


## The language is still moving {#sec-still-moving}

Base R is still absorbing functional idioms, and each addition shrinks the gap between what the language offers and what the lambda calculus requires.

R 4.1 (2021) introduced two changes. The first was `\(x) x + 1` as shorthand for `function(x) x + 1`. Church wrote λ; R now writes `\`, and the one-character form makes anonymous functions cheap enough to use anywhere: inside `lapply()`, inside `Reduce()`, as arguments to `Map()`. Before R 4.1, the eight letters of `function` were a tax on every lambda expression.

The second was the native pipe `|>`. Before R 4.1 the pipe came from the magrittr package (`%>%`), a function that rewrote the call. The native pipe is syntax: `x |> f()` is parsed as `f(x)` before evaluation begins. Function composition is now part of R's grammar.

R 4.4 (2024) added `%||%`, which rlang had provided for years. `x %||% y` returns `x` unless `x` is `NULL`, in which case it returns `y`: a default combinator, the null-coalescing operator of C#, Swift, and Kotlin, and a shorter spelling of `if (is.null(x)) y else x`.

Anonymous functions, composition, and null handling, three operations every functional language has, moved into R's core syntax between 2021 and 2024. The language is still compressing toward its functional core.


## The historical thread {#sec-math-history}

Each idea in this chapter passed through specific hands. The chain is worth tracing because each link solved a problem that the previous link could not.

Could you reduce all of reasoning to calculation? In 1679 the man who had just co-invented calculus, Gottfried Wilhelm Leibniz\index{Leibniz, Gottfried}, imagined a *calculus ratiocinator*, a formal system that would make thinking mechanical. The dream was premature by two centuries. By the 1870s mathematicians had spent decades making their own foundations rigorous, formalizing limits, infinity, and the number line, and a mathematician in Jena saw that logic could be next. Gottlob Frege's\index{Frege, Gottlob} *Begriffsschrift* (1879) introduced quantifiers, variables, and functions as logical primitives, the first formal notation for logic itself. Modern mathematical logic begins there. Church's system of the 1930s kept three constructs, variable, abstraction, and application, and the Church numerals and Church booleans in this chapter are his demonstration that everything else can be built from them.

That was the theory. The question was whether anyone could make it run. McCarthy answered it in 1960 with LISP (@sec-lisp), the first programming language where `(lambda (x) (+ x 1))` was a direct encoding of Church's `λx. x + 1`. But LISP had dynamic scoping, which meant closures did not work the way the theory said they should. Steele and Sussman fixed that in 1975 with Scheme: lexical scoping, first-class closures, and the "Lambda the Ultimate" papers that showed lambda expressions could replace `goto`, assignment, and most control structures.

In the 1970s, a statistician at Bell Labs who wanted to run a regression still had to write a Fortran program, compile it, and wait. S (@sec-s-language) let them type an expression and see the result, with functions and expressions at its core and a computational style borrowed from Fortran. When Ihaka and Gentleman reimplemented S in Auckland in 1993, they gave it Scheme's lexical scoping in place of S's rules (@sec-what-r-inherited). That choice is why closures work in R, why function factories work, and why every Church encoding in this chapter runs without modification.

So the experiment worked. Church showed it was possible, McCarthy made it run, and Ihaka and Gentleman, by choosing Scheme's scoping rules, made it run in R without anyone asking them to.


## References and sources

**Lambda calculus in R:**

- Alonzo Church, "An Unsolvable Problem of Elementary Number Theory" (1936). The paper that started it all.
- Hindley & Seldin, *Lambda-Calculus and Combinators: An Introduction* (2008). Accessible textbook covering Church encoding, beta reduction, fixed points.
- Michaelson, *An Introduction to Functional Programming Through Lambda Calculus* (2011). Gentle path from lambda calculus to practical FP.
- Benjamin Pierce, *Types and Programming Languages* (2002), chapters 5-9. Church booleans, numerals, pairs, recursion via fixed-point combinators.

**Curry-Howard correspondence:**

- Howard, "The Formulae-as-Types Notion of Construction" (1969/1980). Types are propositions, programs are proofs.
- Wadler, "Propositions as Types" (2015, Communications of the ACM). The most readable introduction.

**Category theory connections:**

- Bartosz Milewski, *Category Theory for Programmers* (2019, free online). Accessible introduction with Haskell examples, translatable to R.
- Mac Lane, *Categories for the Working Mathematician* (1971). The standard reference.

**Recreational mathematics in R:**

- [Project Euler](https://projecteuler.net). Mathematical puzzles solvable in any language.
- `Ryacas` package: symbolic math via Yacas CAS.
- `gmp` package: arbitrary precision arithmetic.
- `numbers` package: prime factorization, GCD, modular arithmetic.

**Historical:**

- Leibniz (1679): the dream of a calculus of reasoning.
- Frege (1879): formal logic, *Begriffsschrift*.
- Church (1936): lambda calculus.
- Curry (1934 onward): combinatory logic, Curry-Howard correspondence.
- McCarthy (1960): LISP, first practical lambda calculus.
- Steele & Sussman (1975): Scheme, "Lambda the Ultimate" papers.
- Chambers (1976 onward): S language.
- Ihaka & Gentleman (1993): R, S reimplemented with Scheme's scoping.
